---
title: "Manuscript code"
author: "Jorge Alonso"
date: "2026-06-15"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    code_folding: show
    self_contained: true
---

# 1. Preparing for analysis

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 5,
  dpi = 300)
```

## 1.1. Loading data and packages

```{r, message=FALSE}
# Loading packages
library(tidyverse)
library(treemapify)
library(cowplot)
library(DescTools)
library(scales)
library(DHARMa)
library(ggeffects)
library(lme4)
library(performance)
library(lmtest)
library(lmerTest)
library(emmeans)
```

```{r}
# Cleaning data and memory
rm(list=ls(all=TRUE)) 
gc() 
```

```{r}
# Database link
processed_url <- "https://raw.githubusercontent.com/jmalonso55/Applied-Statistics-for-a-Tropical-Arboretum/refs/heads/main/data/processed_data.csv"
```

```{r}
# Loading data
data_trab <- read.csv(processed_url, sep = ",")

# Checking
summary(data_trab)
```


## 1.2. Subsetting data

```{r}
# Filtering for trees
arvore <- data_trab %>% 
  filter(habito_pad == "Tree")
```

```{r}
# Filtering for Carapa
carapa <- arvore %>%
  filter(genus == "Carapa")
```

```{r}
# Filtering for Arecaceae
palmeira <- data_trab %>% 
  filter(family == "Arecaceae")
```

```{r}
# Filtering for Roystonea
roystonea <- palmeira %>% 
  filter(genus == "Roystonea", !is.na(species))
```

## 1.3. Themes and functions

```{r}
# Frequency table function
freq_tab <- function(data, var, n_classes = 10) {
  data %>%
    mutate(classe = cut({{ var }}, breaks = n_classes, ordered_result = TRUE)) %>%
    count(classe, name = "Freq_obs") %>%
    mutate(Fre_rel = Freq_obs / sum(Freq_obs), Freq_acum_obs = cumsum(Freq_obs), Freq_acum_rel = cumsum(Fre_rel), classe = as.character(classe), lim_inf = as.numeric(sub("^\\((.*),.*\\]$", "\\1", classe)), lim_sup = as.numeric(sub("^\\(.*, *(.*)\\]$", "\\1", classe)), centro_classe = (lim_inf + lim_sup) / 2) %>%
    bind_rows(summarise(., classe = "Total", Freq_obs = sum(Freq_obs), Fre_rel = sum(Fre_rel), Freq_acum_obs = max(Freq_acum_obs), Freq_acum_rel = 1, lim_inf = NA_real_, lim_sup = NA_real_, centro_classe = NA_real_))
}
```

```{r}
# Histogram function
hist_dens <- function(data, var, freq, xlab = NULL, fill = "#A8D0E6", color = "black", alpha = 0.5) {
  freq_plot <- freq %>% filter(classe != "Total")
  breaks_hist <- c(freq_plot$lim_inf, last(freq_plot$lim_sup))
  
  ggplot(data, aes(x = {{ var }})) +
    geom_histogram(breaks = breaks_hist, fill = fill, color = color, alpha = alpha, width = 0.5) +
    labs(x = xlab, y = "Frequency") +
    scale_y_continuous(expand = c(0, 0)) +
    scale_x_continuous(breaks = breaks_hist, labels = scales::label_number(accuracy = 1)) +
    theme_classic() +
    theme(panel.grid.major.y = element_line(color = "grey90", linetype = "dashed"),
          panel.background = element_rect(fill = "grey95"),
          axis.text.x = element_text(angle = 45, hjust = 1))
}
```

# 2. Descriptive analysis

## 2.1. Most frequent families and genera

```{r}
# Filtering top 10 families
familias_top10 <- data_trab %>%
  filter(!is.na(family), family != "") %>%
  count(family, name = "n_family") %>%
  slice_max(n_family, n = 10, with_ties = FALSE) %>%
  pull(family)
```

```{r}
# Abundance of individuals by genus within the top 10 families
abundancia_generos <- data_trab %>%
  filter(family %in% familias_top10, !is.na(genus), genus != "") %>%
  count(family, genus, name = "abundancia") %>%
  group_by(family) %>%
  mutate(rank_genus = rank(-abundancia, ties.method = "first"), label_genus = if_else(rank_genus <= 6, genus, "")) %>%
  ungroup() %>%
  arrange(family, desc(abundancia))
```

```{r}
# Summarizing abundance by family
abundancia_generos %>% 
  group_by(family) %>% 
  summarise(abundancia = sum(abundancia)) %>% 
  arrange(desc(abundancia))
```

```{r}
# Filtering the top 5 genera per family
abundancia_generos %>% 
  filter(rank_genus <= 5) %>% 
  select(-label_genus, -rank_genus)
```

```{r}
# Table visualization
trmp <- 
ggplot(abundancia_generos, aes(area = abundancia, fill = family, subgroup = family)) +
  geom_treemap(color = "white") +
  geom_treemap_subgroup_border(color = "black") +
  geom_treemap_subgroup_text(place = "centre", grow = TRUE, alpha = 0.7, color = "black", fontface = "bold", min.size = 10) +
  geom_treemap_text(aes(label = label_genus), color = "black", place = "bottomright", reflow = TRUE, grow = F, min.size = 4, fontface = "italic", size = 7) +
  theme(legend.position = "none")

trmp
```

```{r}
# Saving
ggsave("treemap.png", plot = trmp, width = 16, height = 9, units = "cm", dpi = 1200)
```

```{r}
# Most abundant species per family
tab_familias <- data_trab %>%
  filter(!is.na(family), family != "", !is.na(name), name != "") %>%
  add_count(family, name = "family_abundance") %>%
  count(family, family_abundance, name, name = "species_abundance") %>%
  group_by(family, family_abundance) %>%
  slice_max(species_abundance, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  slice_max(family_abundance, n = 10, with_ties = FALSE) %>%
  arrange(desc(family_abundance)) %>%
  rename(most_abundant_species = name)

tab_familias
```

```{r, fig.width=8, fig.height=6}
# Table for plot
tab_plot <- tab_familias %>%
  arrange(family_abundance) %>%
  mutate(y = row_number() * 1.4, label_species = paste0(most_abundant_species, " (n = ", species_abundance, ")"))

# Plotting
ggplot(tab_plot) +
  geom_rect(aes(xmin = 0, xmax = family_abundance, ymin = y - 0.30, ymax = y + 0.30), fill = "grey85", color = NA, alpha = 0.5) +
  geom_rect(aes(xmin = 0, xmax = species_abundance, ymin = y - 0.18, ymax = y + 0.18), fill = "#A8D0E6", color = NA) +
  geom_text(aes(x = family_abundance, y = y, label = family_abundance), hjust = -0.25, size = 3.5, color = "grey50") +
  geom_text(aes(x = 10, y = y + 0.43, label = label_species), hjust = 0, size = 2.8, fontface = "bold.italic", color = "#A8D0E6") +
  scale_y_continuous(breaks = tab_plot$y, labels = tab_plot$family, expand = expansion(mult = c(0.04, 0.08))) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.12))) +
  labs(x = "Abundance", y = NULL) +
  coord_cartesian(clip = "off") +
  theme_classic() +
  theme(panel.grid.major.x = element_line(color = "grey90", linetype = "dashed"),
        panel.background = element_rect(fill = "grey98"))
```

## 2.2. Height and DBH distribuition

### 2.2.1. Height for trees

```{r}
# Frequency table for tree height
arv_alt <- freq_tab(arvore, altura)

arv_alt
```

```{r}
# histogram for tree Height
g_arv_alt <- hist_dens(arvore, altura, arv_alt, xlab = "Height (m)")

g_arv_alt
```

### 2.2.2. DBH for trees

```{r}
# Frequency table for tree DBH
arv_dap <- freq_tab(arvore, dap)

arv_dap
```

```{r}
# Histogram for tree DBH
g_arv_dap <- hist_dens(arvore, dap, arv_dap, xlab = "DBH (cm)", fill = "#E6A8A8")

g_arv_dap
```

### 2.2.3. Height for Arecaceae

```{r}
# Frequency table for palm height
palm_alt <- freq_tab(palmeira, altura)

palm_alt
```

```{r}
# Table visualization
g_palm_alt <- hist_dens(arvore, altura, palm_alt, xlab = "Height (m)")

g_palm_alt
```

### 2.2.4. DBH for Arecaceae

```{r}
# Frequency table for palm DBH
palm_dap <- freq_tab(palmeira, dap)

palm_dap
```

```{r}
# Table visualization
g_palm_dap <- hist_dens(palmeira, dap, palm_dap, xlab = "DBH (cm)", fill = "#E6A8A8")

g_palm_dap
```

```{r}
# Investigating DBH distribution by Arecaceae genus to evaluate the pattern observed in the histogram
palmeira %>%
  filter(!is.na(genus)) %>%
  add_count(genus, name = "n_genus") %>%
  filter(n_genus >= sort(unique(n_genus), decreasing = TRUE)[10]) %>%
  mutate(genus = reorder(genus, -dap, median, decreasing = TRUE)) %>%
  ggplot(aes(x = genus, y = dap)) +
  geom_boxplot(fill = "#E6A8A8", alpha = 0.5, outlier.alpha = 0.3) +
  labs(x = "Genus", y = "DBH (cm)") +
  scale_y_continuous(breaks = seq(0, 90, by = 10)) +
  coord_flip() +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        panel.grid.major.x = element_line(color = "grey90", linetype = "dashed"),
        panel.background = element_rect(fill = "grey95"))
```

```{r}
# Median and mean for all palms
palmeira %>% 
  mutate(grupo = if_else(!is.na(genus) & genus == "Roystonea", "Roystonea", "Other genera")) %>% 
  group_by(grupo) %>% 
  summarise(Median = median(dap), Mean = mean(dap), .groups = "drop")
```

### 2.2.5. Joining plots

```{r}
# Setting axis titles
g_arv_alt  <- g_arv_alt  + labs(x = NULL)
g_arv_dap  <- g_arv_dap  + labs(x = NULL, y = NULL)
g_palm_alt <- g_palm_alt
g_palm_dap <- g_palm_dap + labs(y = NULL)
```

```{r}
# Joining plots
linha_arv <- plot_grid(g_arv_alt, g_arv_dap, ncol = 2, align = "hv")
linha_palm <- plot_grid(g_palm_alt, g_palm_dap, ncol = 2, align = "hv")

g_hist <- plot_grid(
  ggdraw() + draw_label("Histograms of trees (n = 3011)", fontface = "bold", size = 14),
  linha_arv,
  ggdraw() + draw_label("Histograms of palms (n = 1362)", fontface = "bold", size = 14),
  linha_palm,
  ncol = 1,
  rel_heights = c(0.10, 1, 0.10, 1)
)

g_hist

```

```{r}
# Saving figure
ggsave("histograms_trees_palms.png", plot = g_hist, width = 20, height = 15, units = "cm", dpi = 1200)
```

## 2.3. Abudance x total basal area

### 2.3.1. For trees

```{r}
# Summary table for trees
tab_res_arv <- arvore %>%
  filter(!is.na(family)) %>%
  group_by(family) %>%
  summarise(n = n(), altura_media = mean(altura), altura_max = max(altura), area_basal_total = sum(area_basal), area_basal_media = mean(area_basal)) %>%
  arrange(desc(area_basal_total))

tab_res_arv
```

```{r}
# Ranking abundance and total basal area
rank_arv <- tab_res_arv %>%
  filter(n >= 10) %>%
  mutate(rank_abundancia = rank(-n, ties.method = "first"), rank_area_basal = rank(-area_basal_total, ties.method = "first"))

rank_arv
```

```{r}
# Converting table to long format
rank_arv_long <- rank_arv %>%
  select(family, rank_abundancia, rank_area_basal) %>%
  pivot_longer(cols = c(rank_abundancia, rank_area_basal), names_to = "variavel", values_to = "ranking") %>%
  mutate(variavel = recode(variavel, rank_abundancia = "Abundance", rank_area_basal = "Total basal area"),
    variavel = factor(variavel, levels = c("Abundance", "Total basal area")))
```

```{r}
# Setting color palette
familias <- sort(unique(rank_arv_long$family))

cores_familias <- colorRampPalette(RColorBrewer::brewer.pal(8, "Dark2"))(length(familias))

names(cores_familias) <- familias
```

```{r}
# Visualizing
bump_arv <- 
ggplot(rank_arv_long, aes(x = variavel, y = ranking, group = family, color = family)) +
  geom_line(alpha = 0.6) +
  geom_point(size = 2) +
  geom_text(data = rank_arv_long %>% filter(variavel == "Abundance"), aes(label = family), hjust = 1.15, size = 3) +
  geom_text(data = rank_arv_long %>% filter(variavel == "Total basal area"), aes(label = family), hjust = -0.15, size = 3) +
  scale_y_reverse(breaks = seq(1, max(rank_arv_long$ranking), by = 1)) +
  scale_color_manual(values = cores_familias) +
  coord_cartesian(clip = "off") +
  labs(x = NULL, y = "Ranking", title = "(A)") +
  theme_classic() +
  theme(text = element_text(family = "serif"),
        legend.position = "none",
        plot.title = element_text(hjust = -0.04, vjust = -6, face = "bold", size = 15),
        plot.title.position = "plot",
        plot.margin = margin(1, 10, 5, 25))

bump_arv
```

### 2.3.2. For palms

```{r}
# Summary table for palm
tab_res_palm <- palmeira %>%
  filter(!is.na(genus)) %>%
  group_by(genus) %>%
  summarise(n = n(), altura_media = mean(altura), altura_max = max(altura), area_basal_total = sum(area_basal), area_basal_media = mean(area_basal)) %>%
  arrange(desc(area_basal_total))

tab_res_palm
```

```{r}
# Ranking abundance and total basal area
rank_palm <- tab_res_palm %>%
  filter(n >= 5) %>%
  mutate(rank_abundancia = rank(-n, ties.method = "first"), rank_area_basal = rank(-area_basal_total, ties.method = "first"))

rank_palm
```

```{r}
# Converting table to long
rank_palm_long <- rank_palm %>%
  select(genus, rank_abundancia, rank_area_basal) %>%
  pivot_longer(cols = c(rank_abundancia, rank_area_basal), names_to = "variavel", values_to = "ranking") %>%
  mutate(variavel = recode(variavel, rank_abundancia = "Abundance", rank_area_basal = "Total basal area"),
    variavel = factor(variavel, levels = c("Abundance", "Total basal area")))
```

```{r}
# Setting color palette
generos <- sort(unique(rank_palm_long$genus))

cores_genus <- colorRampPalette(RColorBrewer::brewer.pal(8, "Dark2"))(length(generos))

names(cores_genus) <- generos
```

```{r}
# Visualizing
bump_palm <- 
ggplot(rank_palm_long, aes(x = variavel, y = ranking, group = genus, color = genus)) +
  geom_line(alpha = 0.6) +
  geom_point(size = 2) +
  geom_text(data = rank_palm_long %>% filter(variavel == "Abundance"), aes(label = genus), hjust = 1.15, size = 3) +
  geom_text(data = rank_palm_long %>% filter(variavel == "Total basal area"), aes(label = genus), hjust = -0.15, size = 3) +
  scale_y_reverse(breaks = seq(1, max(rank_palm_long$ranking), by = 1)) +
  scale_color_manual(values = cores_genus) +
  coord_cartesian(clip = "off") +
  labs(x = NULL, y = "Ranking", title = "(B)") +
  theme_classic() +
  theme(text = element_text(family = "serif"),
        legend.position = "none",
        plot.title = element_text(hjust = -0.04, vjust = -6, face = "bold", size = 15),
        plot.title.position = "plot",
        plot.margin = margin(1, 10, 5, 25))

bump_palm
```

### 2.3.3. Joining plots

```{r, fig.width=10.5, fig.height=6}
# Joining plots
bump_chart <- plot_grid(bump_arv, bump_palm)

bump_chart
```

```{r}
# Saving figure
ggsave("bump_trees_palms.png", plot = bump_chart, width = 26, height = 15, units = "cm", dpi = 600)
```

```{r}
# Table for plotting
tab_dom <- data_trab %>%
  filter(!is.na(family), family != "", !is.na(name), name != "", !is.na(area_basal)) %>%
  count(family, wt = area_basal, name = "family_dominance") %>%
  left_join(data_trab %>%
              filter(!is.na(family), family != "", !is.na(name), name != "", !is.na(area_basal)) %>%
              count(family, name, wt = area_basal, name = "species_dominance") %>%
              group_by(family) %>%
              slice_max(species_dominance, n = 1, with_ties = FALSE) %>%
              ungroup(),
            by = "family") %>%
  slice_max(family_dominance, n = 10, with_ties = FALSE) %>%
  arrange(desc(family_dominance)) %>%
  rename(most_dominant_species = name) %>%
  mutate(prop_dominance = species_dominance / family_dominance)

tab_dom
```

```{r}
# Plotting the top species percentage of total basal area by family
tab_dom %>%
  mutate(family = reorder(family, prop_dominance),
         label_species = paste0(most_dominant_species, " (", round(prop_dominance * 100, 1), "%)"),
         label_total = paste0(round(family_dominance, 1), " m²")) %>%
  ggplot(aes(y = family)) +
  geom_col(aes(x = 1), fill = "grey85", color = "black", width = 0.75) +
  geom_col(aes(x = prop_dominance), fill = "#A8D0E6", color = "black", width = 0.45) +
  geom_text(aes(x = 1, label = label_total), hjust = -0.15, size = 3) +
  geom_text(aes(x = prop_dominance, label = label_species), hjust = -0.05, size = 2.8, fontface = "italic") +
  scale_x_continuous(labels = scales::percent_format(accuracy = 1),
                     limits = c(0, 1.10)) +
  labs(x = "Proportion of family dominance", y = NULL) +
  coord_cartesian(clip = "off") +
  theme_classic() +
  theme(panel.grid.major.x = element_line(color = "grey90", linetype = "dashed"),
        panel.background = element_rect(fill = "grey95"),
        plot.margin = margin(5.5, 80, 5.5, 5.5))
```

# 3. Association between variables

## 3.1. Botanical families x multistems

```{r}
# Contingency table
tab_familia <- table(data_trab$family, data_trab$multicaule_l)

# Checking
tab_familia
```

```{r}
# Contingency table for the 10 most frequent families
filtro_familia <- unique(tab_familias$family)

# Filtering
familia_filtro <- data_trab %>% 
  filter(family %in% filtro_familia)

# Table
familia_10 <- 
table(familia_filtro$family, familia_filtro$multicaule_l)

# Checking
addmargins(familia_10)
```

```{r}
# Proportions by family
prop_fam <- prop.table(familia_10, margin = 1)

round(prop_fam, 2)
```

```{r}
# Chi-squared test
teste_fam <- chisq.test(familia_10)

teste_fam
```


```{r}
# Expected values
round(addmargins(teste_fam$expected), 0)
```

```{r}
# Standardized residuals
round(teste_fam$stdres, 2)
```

```{r}
# Contribution to chi-square
contrib_fam <- teste_fam$residuals^2
round(contrib_fam, 2)
```

```{r}
# Pearson coefficient
coef_cont <- ContCoef(familia_10)

coef_cont
```

```{r}
# Cramer's V coefficient
CramerV(familia_10)
```

```{r}
# Visualizing the table
mosaicplot(familia_10, main = "C = 0.19, p < 0.05", color = c("darkred", "darkblue"), las = 1, cex.axis = 0.6, ylab = "Family", xlab = "Stem condition", dir = c("h", "v"), off = c(15, 3))
```

```{r}
# Preparing data from the contingency table
multicaule_fam <- as.data.frame(familia_10)

names(multicaule_fam) <- c("family", "stem_condition", "n")

multicaule_fam <- multicaule_fam %>%
  group_by(family) %>%
  mutate(total = sum(n), prop = 100 * n / total, prop_yes = prop[stem_condition == "Yes"]) %>%
  ungroup() %>%
  mutate(family = forcats::fct_reorder(family, prop_yes), stem_condition = factor(stem_condition, levels = c("No", "Yes")))

multicaule_fam
```

```{r}
# Plot
contingency <- 
ggplot(multicaule_fam, aes(y = family, x = prop, fill = stem_condition)) +
  geom_col(position = "stack", alpha = 0.8) +
  geom_text(aes(label = ifelse(prop >= 5, paste0(round(prop, 1), "%"), "")), position = position_stack(vjust = 0.5), size = 3.5, color = "grey90") +
  geom_vline(aes(xintercept = 13), linetype = "dashed", color = "grey80", linewidth = 1, alpha = 0.4) +
  scale_fill_manual(values = c("darkred", "darkblue"), labels = c("No", "Yes"),  guide = guide_legend(title = "Multi-stemmed:", reverse = TRUE)) +
  scale_x_continuous(breaks = c(5, 13, 20, 30, 40, 50, 75, 100)) +
  labs(y = "Family", x = "Individuals (%)", caption = paste0("χ² = ", round(teste_fam$statistic, 2), "; df = ", teste_fam$parameter, "; p < 0.001; C = ", round(coef_cont, 2))) +
  theme_classic() +
  theme(legend.position = "bottom",
        plot.caption = element_text(hjust = -0.2, vjust = 13.5, size = 9))

contingency
```

```{r}
# Saving figure
ggsave("multistem.png", plot = contingency, width = 20, height = 13, units = "cm", dpi = 1200)
```

## 3.2. Binomial model for multicaule in Fabaceae

```{r}
# Filtering the database
fabaceae <- arvore %>%
  filter(family == "Fabaceae", !is.na(genus), genus != "") %>%
  add_count(genus, name = "n_genus") %>%
  arrange(desc(n_genus), genus) %>%
  mutate(rank_genus = match(genus, unique(genus))) %>%
  filter(rank_genus <= 10) %>%
  mutate(genus = as.factor(genus)) %>%
  dplyr::select(-n_genus, -rank_genus) %>%
  droplevels()
```

```{r}
# Binomial model
mod_binom <- glm(multicaule ~ altura + genus, data = fabaceae, family = "binomial")
```

```{r, fig.height=5.5, fig.width=9}
# Analysis of residuals
res_mod_binom <- simulateResiduals(mod_binom, n = 1000)
plot(res_mod_binom)
```

```{r}
# Model analysis of deviance
anova(mod_binom, test = "Chisq")
```

```{r}
# Single term analysis of deviance
drop1(mod_binom, test = "Chisq")
```

```{r}
# Model summary
summary(mod_binom)
```

```{r}
# Predição média da probabilidade de multicaule em função da altura
pred_altura <- ggpredict(mod_binom, terms = "altura [all]", margin = "empirical")

# Visualizing
binom_alt <- 
ggplot() +
  geom_jitter(data = fabaceae, aes(x = altura, y = multicaule), height = 0.05, width = 0.05, alpha = 0.35) +
  geom_ribbon(data = pred_altura, aes(x = x, ymin = conf.low, ymax = conf.high), alpha = 0.2) +
  geom_line(data = pred_altura, aes(x = x, y = predicted), linewidth = 1.2) +
  scale_y_continuous(limits = c(-0.1, 1.1), breaks = seq(0, 1, 0.2)) +
  scale_x_continuous(breaks = seq(0, 30, 5)) +
  labs(x = "Height (m)", y = "Predicted probability of multiple stems", title = "(A)") +
  theme_classic() +
  theme(panel.grid.major.x = element_line(color = "grey90", linetype = "dashed"),
        panel.background = element_rect(fill = "grey97"),
        plot.title = element_text(hjust = -0.01, vjust = -6, face = "bold", size = 15),
        plot.title.position = "plot",
        plot.margin = margin(1, 5, 1, 15))

binom_alt
```

```{r}
# Predicting considering genus
pred_genus <- ggpredict(mod_binom, terms = "genus")

# Visualizing
binom_spe <- 
ggplot(pred_genus, aes(x = reorder(x, predicted), y = predicted)) +
  geom_pointrange(aes(ymin = conf.low, ymax = conf.high)) +
  coord_flip() +
  scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.2)) +
  labs(x = "Genus", y = "Predicted probability of multiple stems", title = "(B)") +
  theme_classic() +
  theme(panel.grid.major.x = element_line(color = "grey90", linetype = "dashed"),
        panel.background = element_rect(fill = "grey97"),
        plot.title = element_text(hjust = -0.01, vjust = -6, face = "bold", size = 15),
        plot.title.position = "plot",
        plot.margin = margin(1, 5, 1, 15))

binom_spe
```

```{r, fig.height=4, fig.width=10}
# Joining plots
binom_fab <- plot_grid(binom_alt, binom_spe)

binom_fab
```

```{r}
# Saving figure
ggsave("binomial.png", plot = binom_fab, width = 27, height = 12, units = "cm", dpi = 600)
```

## 3.3. Scatterplot of Height and DBH

```{r}
# Setting interest groups 
data_trab <- data_trab %>%
  mutate(grupo_modelo = case_when(
    genus == "Roystonea" ~ "Roystonea",
    genus == "Carapa" ~ "Carapa",
    TRUE ~ "Other genera"),
    grupo_modelo = factor(grupo_modelo, levels = c("Carapa", "Roystonea", "Other genera")))

# Checking
data_trab %>%
  group_by(grupo_modelo) %>% 
  summarise(n = n(), alt = mean(altura), dap = mean(dap)) %>% 
   mutate(prop = n / sum(n))
```

```{r}
# Scatterplot of height x DBH
scatter <- 
ggplot() +
  geom_point(data = data_trab %>% filter(grupo_modelo == "Other genera"), aes(x = dap, y = altura, color = grupo_modelo), alpha = 0.2, size = 1.2) +
  geom_point(data = data_trab %>% filter(grupo_modelo != "Other genera"), aes(x = dap, y = altura, color = grupo_modelo), alpha = 0.7, size = 1.7) +
  labs(x = "DBH (cm)", y = "Height (m)", color = "Genus:") +
  scale_color_manual(values = c("Carapa" = "darkblue", "Roystonea" = "darkred", "Other genera" = "grey30"), breaks = c("Carapa", "Roystonea", "Other genera")) +
  theme_classic() +
  theme(panel.grid.major.x = element_line(color = "grey90", linetype = "dashed"),
        panel.grid.major.y = element_line(color = "grey90", linetype = "dashed"),
        panel.background = element_rect(fill = "grey97"))

scatter
```

```{r}
# Saving figure
ggsave("scatter.png", plot = scatter, width = 20, height = 12, units = "cm", dpi = 600)
```

# 4. Models for the association between height and diameter

## 4.1. Roystonea

```{r}
# Converting species to factor
roystonea <- roystonea %>%
  mutate(species = factor(species))
```

```{r}
# Summary for Roystonea species
roystonea %>% 
  group_by(species) %>% 
  summarise(n = n(), alt = mean(altura), dap = mean(dap), sle = mean(rel_hd)) %>% 
  mutate(prop = n/sum(n))
```

### 4.1.1. Linear model

```{r}
# Linear model
mod1 <- lm(altura ~ dap + species, data = roystonea)
```

```{r, fig.height=6, fig.width=9}
# Residuals plots
par(mfrow = c(2, 2))
plot(mod1)
```

```{r, fig.height = 5, fig.width = 7}
# Analysis of residuals
res_mod1 <- simulateResiduals(mod1, n = 1000)
plot(res_mod1)
```

### 4.1.2. Log model

```{r}
# Log model
mod_log <- lm(altura ~ log(dap) + species, data = roystonea)
```

```{r, fig.height=6, fig.width=9}
# Residuals plots
par(mfrow = c(2, 2))
plot(mod_log)
```

```{r, fig.height = 5, fig.width = 7}
# Analysis of residuals
res_mod_log <- simulateResiduals(mod_log, n = 1000)
plot(res_mod_log)
```

### 4.1.3. Quadratic model

```{r}
# Quadratic model
mod2 <- lm(altura ~ poly(dap, 2) + species, data = roystonea)
```

```{r, fig.height=6, fig.width=9}
# Residuals plots
par(mfrow = c(2, 2))
plot(mod2)
```

```{r, fig.height = 5, fig.width = 7}
# Analysis of residuals
res_mod2 <- simulateResiduals(mod2, n = 1000)
plot(res_mod2)
```

```{r}
# ANOVA for comparing models
anova(mod1, mod_log, mod2)
```

### 4.1.4. Quadratic with interaction

```{r}
# Quadratic model
mod2_int <- lm(altura ~ species * poly(dap, 2), data = roystonea)
```

```{r, fig.height=6, fig.width=9}
# Residuals plots
par(mfrow = c(2, 2))
plot(mod2_int)
```

```{r, fig.height = 5, fig.width = 7}
# Analysis of residuals
res_mod2_int <- simulateResiduals(mod2_int, n = 1000)
plot(res_mod2_int)
```

```{r}
# Comparing model with and without interaction
anova(mod2, mod2_int)
```

```{r}
# Interaction model summary
summary(mod2_int)
```

```{r}
# Interaction model ANOVA
anova(mod2_int)
```

### 4.1.5. Results

```{r}
# Means comparison between species
emm_sp2 <- emmeans(mod2_int, ~ species | poly(dap, 2), at = list(dap = mean(roystonea$dap, na.rm = TRUE)))

# Multicomparison
multcomp::cld(emm_sp2, adjust = "sidak", Letters = letters, decreasing = T)
```

```{r}
# Model equations

# Equivalent raw polynomial model used only to extract interpretable equations
mod2_int_raw <- lm(altura ~ species * (dap + I(dap^2)), data = roystonea)

# Checking equivalence between predictions
max(abs(predict(mod2_int, newdata = roystonea) - predict(mod2_int_raw, newdata = roystonea)))
```


```{r}
# Extracting coefficients
b <- coef(mod2_int_raw)

# Equations by species
eq_species <- tibble(species = c("borinquena", "oleracea", "regia"),
  intercept = c(
    b["(Intercept)"],
    b["(Intercept)"] + b["speciesoleracea"],
    b["(Intercept)"] + b["speciesregia"]),
  
  beta_dap = c(
    b["dap"],
    b["dap"] + b["speciesoleracea:dap"],
    b["dap"] + b["speciesregia:dap"]),
  
  beta_dap2 = c(
    b["I(dap^2)"],
    b["I(dap^2)"] + b["speciesoleracea:I(dap^2)"],
    b["I(dap^2)"] + b["speciesregia:I(dap^2)"])) %>%
  
  mutate(equation = paste0("Height = ", round(intercept, 2), ifelse(beta_dap >= 0, " + ", " - "), abs(round(beta_dap, 3)), " DBH", ifelse(beta_dap2 >= 0, " + ", " - "), abs(round(beta_dap2, 5)), " DBH²"))

# Checking
eq_species %>%
  select(species, equation)
```


```{r}
# Generating dataframe with values to predict
pred_species <- roystonea %>%
  group_by(species) %>%
  summarise(dap_min = min(dap), dap_max = max(dap), .groups = "drop") %>%
  group_by(species) %>%
  reframe(dap = seq(dap_min, dap_max, length.out = 100))
```

```{r}
# Predicting within the DBH range observed for each species
pred_fit <- predict(mod2_int, newdata = pred_species, interval = "confidence")

pred_species <- pred_species %>%
  mutate(altura_pred = pred_fit[, "fit"], altura_lwr  = pred_fit[, "lwr"], altura_upr  = pred_fit[, "upr"])
```

```{r}
# Species labels for legend
species_labels <- c(borinquena = "italic('Roystonea borinquena')", oleracea   = "italic('Roystonea oleracea')", regia      = "italic('Roystonea regia')")

# Plotting the results
roys <- 
ggplot(roystonea, aes(x = dap, y = altura, color = species)) +
  geom_point(alpha = 0.45, size = 1.5) +
  geom_ribbon(data = pred_species, aes(x = dap, ymin = altura_lwr, ymax = altura_upr, fill = species, group = species), alpha = 0.15, color = NA, inherit.aes = FALSE) +
  geom_line(data = pred_species, aes(x = dap, y = altura_pred, color = species, group = species), linewidth = 1) +
  scale_color_discrete(breaks = names(species_labels), labels = parse(text = species_labels)) +
  scale_x_continuous(breaks = seq(10, 90, 10)) +
  scale_y_continuous(breaks = seq(0, 45, 7)) +
  labs(x = "DBH (cm)", y = "Height (m)", color = "Species", fill = "Species") +
  guides(fill = "none") +
  theme_classic() +
  theme(legend.position = "bottom",
        panel.grid.major.x = element_line(color = "grey90", linetype = "dashed"),
        panel.grid.major.y = element_line(color = "grey90", linetype = "dashed"),
        panel.background = element_rect(fill = "grey97"))

roys
```

```{r}
# Saving figure
ggsave("roystonea.png", plot = roys, width = 20, height = 12, units = "cm", dpi = 600)
```


## 4.2. Carapa

### 4.2.1. Linear model

```{r}
# Linear model
mod1c <- lm(altura ~ dap, data = carapa)
```


```{r, fig.height=6, fig.width=9}
# Residuals plots
par(mfrow = c(2, 2))
plot(mod1c)
```


```{r, fig.height = 5, fig.width = 7}
# Analysis of residuals
res_mod1c <- simulateResiduals(mod1c, n = 1000)
plot(res_mod1c)
```

### 2.2.2. Log model

```{r}
# Log model
mod_logc <- lm(altura ~ log(dap), data = carapa)
```


```{r, fig.height=6, fig.width=9}
# Residuals plots
par(mfrow = c(2, 2))
plot(mod_logc)
```


```{r, fig.height = 5, fig.width = 7}
# Analysis of residuals
res_mod_logc <- simulateResiduals(mod_logc, n = 1000)
plot(res_mod_logc)
```

```{r}
# log model ANOVA
anova(mod_logc)
```

```{r}
# log model summary
summary(mod_logc)
```


### 2.2.3. Quadratic model

```{r}
# Quadratic model
mod2c <- lm(altura ~ dap + I(dap^2), data = carapa)
```


```{r, fig.height=6, fig.width=9}
# Residuals plots
par(mfrow = c(2, 2))
plot(mod2c)
```


```{r, fig.height = 5, fig.width = 7}
# Analysis of residuals
res_mod2c <- simulateResiduals(mod2c, n = 1000)
plot(res_mod2c)
```

```{r}
# Model comparison
AIC(mod1c, mod_logc, mod2c)
BIC(mod1c, mod_logc, mod2c)
```

### 2.2.4. Mixed effects log model

```{r}
# Filtering for beds with three or more trees
carapa_f <- carapa %>% 
  add_count(sec_cant, name = "n_sec_cant") %>% 
  filter(n_sec_cant >= 3) %>% 
  mutate(sec_cant = factor(sec_cant))
```


```{r}
# Summary per bed
carapa_f %>% 
  group_by(sec_cant) %>% 
  summarise(n = n(), alt = mean(altura), dap = mean(dap), sle = mean(rel_hd)) %>% 
  mutate(prop = n/sum(n)) %>% 
  arrange(desc(n))
```


```{r}
# log model with beds random effects
mod1m <- lmer(altura ~ log(dap) + (1 | sec_cant), data = carapa_f, REML = F)
```


```{r, fig.height=5, fig.width=7}
# Analysis of residuals
res_mod1m <- simulateResiduals(mod1m, n = 1000)
plot(res_mod1m)
```


```{r}
# Checking random effects residual
plotResiduals(res_mod1m, form = carapa_f$sec_cant)
```


```{r}
# Refitting model for comparison
mod_logc_f <- lm(altura ~ log(dap), data = carapa_f)
```


```{r}
# Comparing models without and with random effects
AIC(mod_logc_f, mod1m)
BIC(mod_logc_f, mod1m)
```

```{r}
# log model R²
r2(mod_logc)
```

```{r}
# mixed model R²
r2(mod1m)
```

```{r}
# Checking for random effects significance
ranova(mod1m)
```


### 2.2.5. Mixed effects quadratic model

```{r}
# Quadratic model with beds random effects
mod2m <- lmer(altura ~ poly(dap, 2) + (1 | sec_cant), data = carapa_f, REML = F)
```

```{r, fig.height=5, fig.width=7}
# Analysis of residuals
res_mod2m <- simulateResiduals(mod2m, n = 1000)
plot(res_mod2m)
```


```{r}
# Checking random effects residual
plotResiduals(res_mod2m, form = carapa_f$sec_cant)
```


```{r}
# Comparing the best models
AIC(mod_logc_f, mod1m, mod2m)
BIC(mod_logc_f, mod1m, mod2m)
```


```{r}
# log model R²
r2(mod_logc_f)
```


```{r}
# Quadratic model R²
r2(mod2m)
```

```{r}
# Random effects significance
ranova(mod2m)
```


```{r}
# Refitting with REML = T
mod2m <- lmer(altura ~ poly(dap, 2) + (1 | sec_cant), data = carapa_f, REML = T)
```

```{r}
# Mixed model summary
summary(mod2m)
```

```{r}
# Mixed model summary
anova(mod2m)
```


```{r}
# Mixed model summary
summary(mod2m)
```

```{r}
# Random effects significance
ranova(mod2m)
```

```{r}
# Variance and correlation components
VarCorr(mod2m)
```


```{r}
# Intraclass Correlation Coefficient
icc(mod2m)
```


### 2.2.6. Results visualization

```{r}
# Model equations

# Equivalent raw polynomial model used only to extract interpretable equations
mod2m_raw <- lmer(altura ~ dap + I(dap^2) + (1 | sec_cant), data = carapa_f, REML = T, control = lmerControl(check.scaleX = "ignore"))

# Checking equivalence between predictions
max(abs(predict(mod2m) - predict(mod2m_raw)))
```


```{r}
# Fixed-effect coefficients
b <- fixef(mod2m_raw)

# Population-level equation
eq_carapa_fixed <- tibble(
  model = "Population-level curve",
  intercept = b["(Intercept)"],
  beta_dap = b["dap"],
  beta_dap2 = b["I(dap^2)"]) %>%
  
  mutate(equation = paste0("Height = ", round(intercept, 2), ifelse(beta_dap >= 0, " + ", " - "), abs(round(beta_dap, 3)), " DBH", ifelse(beta_dap2 >= 0, " + ", " - "), abs(round(beta_dap2, 5)), " DBH²"))

eq_carapa_fixed
```


```{r}
# Random intercepts by arboretum bed
ran_bed <- ranef(mod2m_raw)$sec_cant %>%
  tibble::rownames_to_column("sec_cant") %>%
  rename(random_intercept = `(Intercept)`)

# Bed-specific equations
eq_carapa_bed <- ran_bed %>%
  mutate(intercept = b["(Intercept)"] + random_intercept,
    beta_dap = b["dap"],
    beta_dap2 = b["I(dap^2)"],
    equation = paste0(sec_cant, ": Height = ",
      round(intercept, 2),
      ifelse(beta_dap >= 0, " + ", " - "),
      abs(round(beta_dap, 3)), " DBH",
      ifelse(beta_dap2 >= 0, " + ", " - "),
      abs(round(beta_dap2, 5)), " DBH²"))

# Checking
eq_carapa_bed %>%
  select(sec_cant, random_intercept, equation) %>% 
  arrange(desc(random_intercept))
```




```{r}
# Predicting for plotting
pred_mod2m <- ggpredict(mod2m, terms = "dap [all]", type = "fixed")

# Visualizing
ggplot(pred_mod2m, aes(x = x, y = predicted)) +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.2) +
  geom_line(linewidth = 1.2) +
  geom_point(data = carapa_f, aes(x = dap, y = altura), inherit.aes = FALSE, alpha = 0.7) +
  labs(x = "DAP", y = "Altura predita", title = "") +
  theme_classic()
```

```{r}
# Random effects of arboretum beds
efeito_cant <- ranef(mod2m)$sec_cant %>%
  as.data.frame() %>%
  rownames_to_column("sec_cant") %>%
  rename(efeito_canteiro = `(Intercept)`)

# DBH sequence for prediction
dap_seq <- seq(min(carapa_f$dap, na.rm = TRUE), max(carapa_f$dap, na.rm = TRUE), length.out = 300)
```

```{r}
# Global mean curve
pred_global <- tibble(dap = dap_seq) %>%
  mutate(altura_pred = predict(mod2m, newdata = ., re.form = NA))
```

```{r}
# Predicted curves by arboretum bed, using only observed DBH range in each bed
pred_cant <- carapa_f %>%
  group_by(sec_cant) %>%
  group_modify(~ tibble(dap = seq(min(.x$dap, na.rm = TRUE), max(.x$dap, na.rm = TRUE), length.out = 300))) %>%
  ungroup() %>%
  mutate(altura_pred = predict(mod2m, newdata = ., re.form = NULL)) %>%
  left_join(efeito_cant, by = "sec_cant")
```


```{r}
# Plot with points, bed-specific curves, and population-level curve
carapa <- 
ggplot(carapa_f, aes(dap, altura)) +
  geom_point(alpha = 0.45, size = 1.5, color = "gray45") +
  geom_line(data = pred_cant, aes(y = altura_pred, group = sec_cant), color = "gray75", linewidth = 0.6, alpha = 0.8) +
  geom_line(data = pred_global, aes(x = dap, y = altura_pred), inherit.aes = FALSE, linewidth = 1.3, color = "gray20") +
  scale_y_continuous(limits = c(0, NA), breaks = seq(0, 25, 5)) +
  scale_x_continuous(limits = c(0, NA), breaks = seq(0, 110, 15)) +
  labs(x = "DBH (cm)", y = "Height (m)", title = "(A)") +
  theme_classic() +
  theme(panel.grid.major.x = element_line(color = "grey90", linetype = "dashed"),
        panel.grid.major.y = element_line(color = "grey90", linetype = "dashed"),
        panel.background = element_rect(fill = "grey97"),
        plot.title = element_text(hjust = -0.02, vjust = -4, face = "bold", size = 15),
        plot.title.position = "plot",
        plot.margin = margin(1, 5, 1, 15))

carapa
```


```{r}
# Extracting random intercepts for arboretum beds
rank_canteiros <- ranef(mod2m)$sec_cant %>%
  as.data.frame() %>%
  rownames_to_column("sec_cant") %>%
  rename(efeito_canteiro = `(Intercept)`) %>%
  arrange(desc(efeito_canteiro)) %>%
  mutate(rank = row_number(), interpretacao = case_when(
      efeito_canteiro > 0 ~ "Acima da média",
      efeito_canteiro < 0 ~ "Abaixo da média",
      TRUE ~ "Na média"))
```


```{r}
# Visualizing
canteiro <- 
ggplot(rank_canteiros, aes(x = reorder(sec_cant, efeito_canteiro), y = efeito_canteiro)) +
  geom_col() +
  geom_hline(yintercept = 0, linetype = "dashed") +
  coord_flip() +
  labs(x = "Arboretum bed", y = "Estimated random intercept", title = "(B)") +
  theme_classic() +
  theme(panel.background = element_rect(fill = "grey97"),
        plot.title = element_text(hjust = -0.02, vjust = -4, face = "bold", size = 15),
        plot.title.position = "plot",
        plot.margin = margin(1, 5, 1, 15))

canteiro
```


```{r, fig.height=6, fig.width=14}
# Joining plots
car_mix <- plot_grid(carapa, canteiro)

car_mix
```


```{r}
# Saving figure
ggsave("carapa.png", plot = car_mix, width = 24, height = 10, units = "cm", dpi = 600)
```
