---
title: "Descriptive analysis"
author: "Jorge Alonso"
date: "2026-05-19"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    code_folding: show
    self_contained: true
---

# 1. Preparing

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 5,
  dpi = 300)
```

## 1.1. Loading packages and data

```{r, message = FALSE}
# Loading packages
library(tidyverse)
library(skimr)
library(stringr)
library(stringi)
library(DescTools)
library(moments)
library(treemapify)
```


```{r, message=FALSE, warning=FALSE}
# Cleaning memory
rm(list=ls(all=TRUE))
invisible(gc())
```


```{r}
# Database link
raw_url <- "https://raw.githubusercontent.com/jmalonso55/Applied-Statistics-for-a-Tropical-Arboretum/refs/heads/main/data/processed_data.csv"
```


```{r}
# Loading data
data_trab <- read.csv(raw_url, sep = ",")

# Checking
summary(data_trab)
```


## 1.2. Themes and functions

```{r}
# Theme for plots 1
tema <- theme(panel.grid.major.x = element_line(color = "grey90", linetype = "dashed"),
        panel.background = element_rect(fill = "grey95"))
```

```{r}
# Theme for plots 2
tema2 <- theme(panel.grid.major.y = element_line(color = "grey90", linetype = "dashed"),
        panel.background = element_rect(fill = "grey95"))
```

```{r}
# Defining a function for the coefficient of variation
var_coef <- function(x) {
  (sd(x, na.rm = T)/mean(x, na.rm = T)) * 100}
```

# 2. Descriptive statistics

## 2.1. Frequency distribution table

### 2.1.1. Qualitative variable (family)

```{r}
# Creating a frequency table by family
familia <- data_trab %>% 
  filter(!is.na(family)) %>% 
  count(family, name = "Freq_obs") %>% 
  arrange(desc(Freq_obs))

# Checking
head(familia, 5)
```

```{r}
# Calcular a frequência relativa e acumulada
familia <- familia %>%
  mutate(Fre_rel = Freq_obs / sum(Freq_obs),
    Freq_acum_obs = cumsum(Freq_obs),
    Freq_acum_rel = cumsum(Fre_rel))

# Adicionar linha de total
familia <- familia %>%
  bind_rows(summarise(., family = "Total", 
                      Freq_obs = sum(Freq_obs), 
                      Fre_rel = sum(Fre_rel), 
                      Freq_acum_obs = sum(Freq_obs), 
                      Freq_acum_rel = 1))

# Visualizar as 20 primeiras linhas da tabela
head(familia, 10)
```


```{r}
# Table visualization
ggplot(head(familia, 20), aes(x = Freq_obs, y = reorder(family, Freq_obs))) +
  geom_col(fill = "#A8D0E6") +
  labs(y = "Family", x = "Observed frequency") +
  scale_x_continuous(expand = c(0, 0), breaks = c(seq(0, 1400, by = 200))) +
  theme_classic() +
  tema
```


### 2.1.2. Discrete quantitative variable (number of stems)

```{r}
# Creating a frequency table by number of stems
nfuste <- data_trab %>% 
  count(n_caule, name = "Freq_obs") %>% 
  arrange(desc(Freq_obs))

# Checking
head(nfuste, 5)
```


```{r}
# Calcular a frequência relativa e acumulada
nfuste <- nfuste %>%
  mutate(Freq_rel = Freq_obs / sum(Freq_obs),
    Freq_acum_obs = cumsum(Freq_obs),
    Freq_acum_rel = cumsum(Freq_rel),
    n_fuste = as.character(n_caule)) 

# Checking
nfuste
```

```{r}
# Table visualization
nfuste %>%
  mutate(n_fuste = as.integer(as.character(n_fuste)),
    n_fuste = factor(n_fuste, levels = rev(sort(unique(n_fuste))))) %>%
  ggplot(aes(x = Freq_obs, y = n_fuste)) +
  geom_col(fill = "#A8D0E6") +
  labs(y = "Number of stems", x = "Observed frequency") +
  scale_x_continuous(expand = c(0, 0)) +
  theme_classic() +
  tema
```


### 2.1.3. Continuous quantitative variable (slenderness)

```{r}
# Definir classes de altura
data_trab <- data_trab %>%
    mutate(classe_sle = cut(rel_hd, breaks = 10))

# Calcular as frequências para cada classe de dap
tab_sle <- data_trab %>%
  count(classe_sle, name = "Freq_obs") %>%
  mutate(Fre_rel = Freq_obs / sum(Freq_obs),
    Freq_acum_obs = cumsum(Freq_obs),
    Freq_acum_rel = cumsum(Fre_rel))

tab_sle <- tab_sle %>%
  bind_rows(summarise(., classe_sle = "Total",
                      Freq_obs = sum(Freq_obs),
                      Fre_rel = sum(Fre_rel),
                      Freq_acum_obs = sum(Freq_obs),
                      Freq_acum_rel = 1))

# Visualizar a tabela de frequência da temperatura
tab_sle
```

```{r}
# Class limits
limites_sle <- tibble(classe_sle = levels(data_trab$classe_sle)) %>%
  mutate(classe_txt = str_remove_all(classe_sle, "\\(|\\]|\\[")) %>%
  separate(classe_txt, into = c("limite_inf", "limite_sup"), sep = ",", convert = TRUE)

limites_sle
```


```{r}
# Table visualization
  ggplot(data_trab, aes(x = rel_hd)) +
  geom_histogram(bins = 10, fill = "#A8D0E6", color = "black", breaks = limites_sle$limite_inf) +
  labs(x = "Slenderness", y = "Observed frequency") +
  scale_y_continuous(expand = c(0, 0)) +
  scale_x_continuous(breaks = limites_sle$limite_inf) +
  theme_classic() +
  tema2
```


## 2.2. Bivariate analysis

### 2.2.1. Qualitative x quantitative

```{r}
# Mean and sd of height by stem condition
data_trab %>% 
  group_by(multicaule_l) %>% 
  summarise(Mean = mean(altura), Standard_deviation = sd(altura)) %>% 
  arrange(desc(Mean))
```

```{r}
# Visualizing in a plot
  ggplot(data_trab, aes(x = multicaule_l, y = altura)) +
  geom_jitter(height = 0.05, width = 0.1, alpha = 0.4, color = "lightblue") +
  stat_summary(fun.data = mean_sdl, size = 0.5, linewidth = 0.5, fun.args = list(mult = 1), color = "gray35") +
  stat_summary(fun = mean, geom = "text", aes(label = sprintf("%.2f", after_stat((y)))), position = position_nudge(x = 0.25), size = 3, color = "gray25") +
  labs(y = "Height (m)", x = "Bifurcated") +
  scale_y_continuous(breaks = c(0, 5, 10, 15, 20, 25, 30, 35, 40), limits = c(0, NA)) +
  theme_classic() +
  tema2
```


### 2.2.2. Qualitative with qualitative

```{r}
# Contingency table
tab_familia <- table(data_trab$family, data_trab$multicaule_l)

# Observing
tab_familia
```


```{r}
# Contingency table for the 10 most frequent families
filtro_familia <- unique(head(familia$family, 10))

# Filtering
familia_filtro <- data_trab %>% 
  filter(family %in% filtro_familia)

# Creating the table
familia_10 <- 
table(familia_filtro$family, familia_filtro$multicaule_l)

# Observing
addmargins(familia_10)
```

```{r}
# Running the test
teste_fam <- chisq.test(familia_10)

# Inspecting the expected values
round(addmargins(teste_fam$expected), 0)
```

```{r}
# Test result
teste_fam
```


```{r}
# Visualizing the table
mosaicplot(familia_10, main = "", color = c("darkred", "darkblue"), las = 1, cex.axis = 0.6, ylab = "Family", xlab = "Stem condition", dir = c("h", "v"), off = c(15, 3))
```



```{r}
# Contingency coefficient
ContCoef(familia_10)
```

```{r}
# Abundance of individuals by genus within the top 10 families
abundancia_generos <- data_trab %>%
  filter(
    family %in% filtro_familia,
    !is.na(genus),
    genus != ""
  ) %>%
  count(family, genus, name = "abundancia") %>%
  arrange(family, desc(abundancia))

abundancia_generos
```


```{r}
# Table visualization
ggplot(abundancia_generos, aes(area = abundancia, fill = family, subgroup = family, label = genus)
) +
  geom_treemap(color = "white") +
  geom_treemap_subgroup_border(color = "black", linewidth = 0.7) +
  geom_treemap_subgroup_text(place = "centre", grow = TRUE, alpha = 0.6, color = "black", fontface = "bold") +
  geom_treemap_text(color = "black", place = "centre", reflow = TRUE, size = 8) +
  labs(fill = "Family") +
  theme(legend.position = "none",
    plot.title = element_text(size = 14, face = "bold"))
```


### 2.2.3. Association between quantitative variables

```{r}
# Correlation between dendrometric variables
with(data_trab, cor.test(altura, dap, method = "spearman"))
```


```{r}
# Graphical visualization
ggplot(data_trab, aes(x = dap, y = altura)) +
  geom_point(alpha = 0.7, color = "lightblue") +
  labs(x = "DBH (cm)", y = "Height (m)") +
  theme_classic() +
  tema +
  theme(panel.grid.major.y = element_line(color = "grey90", linetype = "dashed"))
```


### 2.2.4.Combining variables to explore the dataset

```{r}
# Alternative graphical visualization
ggplot(data_trab, aes(x = dap, y = altura, color = multicaule_l)) +
  geom_point(alpha = 0.7) +
  labs(x = "DBH (cm)", y = "Height (m)", color = "Multi-stem") +
  theme_classic() +
  tema +
  theme(panel.grid.major.y = element_line(color = "grey90", linetype = "dashed"))
```


```{r}
# Alternative graphical visualization with faceting
ggplot(data_trab, aes(x = dap, y = altura, color = multicaule_l)) +
  geom_point(alpha = 0.7) +
  labs(x = "DBH (cm)", y = "Height (m)", color = "Multi-stem") +
  facet_grid(~multicaule_l) +
  theme_classic() +
  tema +
  theme(panel.grid.major.y = element_line(color = "grey90", linetype = "dashed"))
```

```{r}
# Alternative graphical visualization - slenderness classes
ggplot(data_trab, aes(x = dap, y = altura, color = classe_sle)) +
  geom_point(alpha = 0.7) +
  labs(x = "DBH (cm)", y = "Height (m)", color = "Slenderness class") +
  theme_classic() +
  tema +
  theme(panel.grid.major.y = element_line(color = "grey90", linetype = "dashed"))
```


```{r}
# Alternative graphical visualization with faceting - slenderness classes
ggplot(data_trab, aes(x = dap, y = altura, color = classe_sle)) +
  geom_point(alpha = 0.7) +
  labs(x = "DBH (cm)", y = "Height (m)", color = "Slenderness class") +
  facet_wrap(~classe_sle) +
  theme_classic() +
  tema +
  theme(panel.grid.major.y = element_line(color = "grey90", linetype = "dashed"))
```


## 2.3. Summary Measures

### 2.3.1. Central Tendency

```{r}
# Means of numerical variables
data_trab %>%
  filter(family %in% filtro_familia) %>% 
  group_by(family) %>% 
  summarise(n = n(), across(c(altura, dap, densidade, n_caule, rel_hd), mean)) %>% 
  arrange(desc(altura))
```

```{r}
# Comparing mean and median
data_trab %>%
  filter(family %in% filtro_familia) %>% 
  group_by(family) %>% 
  summarise(n = n(), dbh_mean = mean(dap), dbh_median = median(dap)) %>% 
  arrange(desc(dbh_mean))
```


### 2.3.2. Dispersion

```{r}
# Mean, standard deviation, and coefficient of variation
data_trab %>%
  filter(family %in% filtro_familia) %>% 
  group_by(family) %>% 
  summarise(alt_media = mean(altura), alt_dp = sd(altura), alt_cv = var_coef(altura)) %>% 
  arrange(desc(alt_cv))
```



```{r}
# Selecting the top 10 beds with the most plants
canteiros <- data_trab %>%
  count(sec_cant, sort = TRUE) %>%
  slice_head(n = 10)

canteiros
```


```{r}
# Height mean, standard deviation, and coefficient of variation - beds
data_trab %>%
  filter(sec_cant %in% pull(canteiros, sec_cant)) %>% 
  group_by(sec_cant) %>% 
  summarise(n = n(), alt_media = mean(altura), alt_dp = sd(altura), alt_cv = var_coef(altura)) %>% 
  arrange(desc(alt_cv))
```


```{r}
# DBH mean, standard deviation, and coefficient of variation - beds
data_trab %>%
  filter(sec_cant %in% pull(canteiros, sec_cant)) %>% 
  group_by(sec_cant) %>% 
  summarise(n = n(), dbh_media = mean(dap), dbh_dp = sd(dap), dbh_cv = var_coef(dap)) %>% 
  arrange(desc(dbh_cv))
```


### 2.3.3. Measures of relative position

```{r}
# Quartiles
quantile(data_trab$rel_hd)
```

```{r}
# Deciles
quantile(data_trab$rel_hd, probs = c(seq(0, 1, by = 0.1)))
```

```{r}
# Percentiles
quantile(data_trab$rel_hd, probs = c(0.01, 0.05, 0.95, 0.99))
```


```{r}
# Boxplot
data_trab %>%
  filter(family %in% filtro_familia) %>% 
  ggplot(aes(x = rel_hd, y = reorder(family, rel_hd))) +
  geom_boxplot(alpha = 0.4) +
  labs(y = "Family", x = "Slenderness") +
  theme_classic() +
  tema
```

### 2.3.4. Measures of distribution shape

```{r}
# Skewness
skewness(data_trab$dap)
```


```{r}
# Kurtosis
kurtosis(data_trab$dap)
```


```{r}
# Observed frequency
ggplot(data = data_trab, aes(x = dap)) +
  geom_histogram(bins = 30, fill = "#A8D0E6", color = "black", alpha = 0.4) +
  labs(x = "DBH (cm)", y = "Frequência observada") +
  scale_y_continuous(expand = c(0, 0)) +
  theme_classic() +
  tema2
```


