---
title: "Probability and random variables"
author: "Jorge Alonso"
date: "2026-05-21"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    code_folding: show
    self_contained: true
---

# 1. Preparing, loading packages and data

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 5,
  dpi = 300)
```


```{r, message = FALSE}
# Loading packages
library(tidyverse)
```


```{r, message = FALSE}
# Cleaning memory
rm(list=ls(all=TRUE))
invisible(gc())
```


```{r}
# Database link
processed_url <- "https://raw.githubusercontent.com/jmalonso55/Applied-Statistics-for-a-Tropical-Arboretum/refs/heads/main/data/processed_data.csv"
```


```{r}
# Loading data
data_trab <- read.csv(processed_url, sep = ",")

# Checking
summary(data_trab)
```

```{r}
# Filtering to remove NA from the family column
data_trab <- data_trab %>% 
  filter(!is.na(family))
```


# 2. Empirical probabilities

If an individual is randomly selected from the arboretum, what can we say about the probability of its family, stem status, or DBH class?

```{r}
# Randomly sampling one individual
sample_n(data_trab, 1)
```


```{r}
# If a plant is randomly selected in the arboretum, what is the probability that it is multi-stemmed?
total <- nrow(data_trab)
bif <- nrow(subset(data_trab, multicaule == 1))

p_evA <- bif/total
p_evA
```


```{r}
# Complementary event
p_evAc <- 1 - p_evA

p_evAc
```


```{r}
# Using dplyr
data_trab %>%
  summarise(p_multicaule = mean(multicaule == 1))
```


```{r}
# What is the probability that the individual is a palm?
p_evB <- data_trab %>%
  summarise(p_palmeira = mean(family == "Arecaceae")) %>% 
  pull()

p_evB
```

```{r}
# What is the probability of having DBH greater than or equal to 70 cm?
p_evC <- data_trab %>%
  summarise(p_dap70 = mean(dap >= 70)) %>% 
  pull()

p_evC
```


```{r}
# Intersection event: What is the probability that the individual is a palm and has DBH greater than or equal to 70 cm?
p_B_inter_C <- data_trab %>%
  summarise(prob = mean(family == "Arecaceae" & dap >= 70)) %>% 
  pull()

p_B_inter_C
```


```{r}
# Union event: What is the probability that the individual is a palm or have a DBH greater than or equal to 70 cm?
(p_evB + p_evC) - p_B_inter_C
```

```{r}
# Or:
p_B_union_C <- data_trab %>%
  summarise(prob = mean(family == "Arecaceae" | dap >= 70)) %>% 
  pull()

p_B_union_C
```


```{r}
# Conditional probability: Among Fabaceae individuals, what is the probability of observing multi-stemmed individuals?
data_trab %>%
  filter(family == "Fabaceae") %>%
  summarise(p_bif_LEG = mean(multicaule == 1)) %>% 
  pull()
```

```{r}
# Conditional probability: Among multi-stemmed individuals, what is the probability of observing a plant from the Fabaceae family?
data_trab %>%
  filter(multicaule == 1) %>%
  summarise(p_bif_LEG = mean(family == "Fabaceae", na.rm = TRUE)) %>% 
  pull()
```


# 3. Random variables

## 3.1. Binomial distribution 

```{r}
# Setting the number of trials (n) and the number of successes of interest (x)
n <- 20
x <- 3
```


```{r}
# If 20 plants are randomly selected, what is the probability of observing exactly three multi-stemmed individuals?
dbinom(x, size = n, prob = p_evA)
```


```{r}
# What is the probability of observing three or fewer multi-stemmed plants in a sample of 20 plants?
pbinom(x, size = n, prob = p_evA, lower.tail = TRUE)
```

```{r}
# And what is the probability of observing more than three multi-stemmed plants in a sample of 20 plants?
pbinom(x, size = n, prob = p_evA, lower.tail = FALSE)
```


```{r}
# Plotting the probability distribution

# Defining the sample size and possible outcomes
n <- 20
x <- 0:n

# Calculating binomial probabilities
prob <- dbinom(x, size = n, prob = p_evA)

# Creating the data frame for plotting
dados_binom <- tibble(multicaule = x,  probabilidade = prob)

# Plotting the probability distribution
ggplot(dados_binom, aes(x = multicaule, y = probabilidade)) +
  geom_col(width = 0.3, fill = "lightblue") +
  scale_x_continuous(breaks = seq(0, 20, by = 1)) +
  labs(x = "Number of multi-stemmed individuals in a sample of 20", y = "Probability") +
  theme_classic()
```



```{r}
# Estimating binomial parameters for multi-stemmed in a sample of 20 plants
binom_resumo <- tibble(n = n, p = p_evA, esperanca = n * p, variancia = n * p * (1 - p), desvio_padrao = sqrt(n * p * (1 - p)))

binom_resumo
```


## 3.2. Poisson distribution

Does the count of individuals with high DBH per planting bed behave as a Poisson variable?

```{r}
# Calculating the 95th percentile for DBH
quantile(data_trab$dap, probs = c(0.95))
```

```{r}
# Selecting the 50 beds with the highest number of plants
top_beds <- data_trab %>%
  count(sec_cant, name = "n_total", sort = TRUE) %>%
  slice_head(n = 50)

# Counting plants with DBH above the 95th percentile in each selected bed
bdap_beds <- data_trab %>%
  filter(sec_cant %in% top_beds$sec_cant) %>%
  group_by(sec_cant) %>%
  summarise(n_total = n(), n_dap = sum(dap > 100))

head(bdap_beds)
```


```{r}
# Checking mean, variance, and dispersion
bdap_beds %>%
  summarise(mean_n_dap = mean(n_dap), var_n_dap = var(n_dap), dispersion_ratio = var_n_dap / mean_n_dap)
```


The Poisson distribution can be used as a theoretical reference, but the data show overdispersion. Therefore, the Poisson distribution does not perfectly describe this pattern.


```{r}
# Theoretical Poisson distribution

# Estimating the Poisson parameter
lambda <- mean(bdap_beds$n_dap)

# Generating theoretical Poisson probabilities
poisson_theoretical <- tibble(n_dap = 0:max(bdap_beds$n_dap), probability = dpois(n_dap, lambda = lambda))

poisson_theoretical
```


```{r}
# Plotting the theoretical Poisson distribution
ggplot(poisson_theoretical, aes(x = n_dap, y = probability)) +
  geom_col(width = 0.4, fill = "lightblue") +
  scale_x_continuous(breaks = 0:max(poisson_theoretical$n_dap)) +
  labs(x = "Number of plants with DBH > 100 cm per bed", y = "Probability") +
  theme_classic()
```



## 3.3. Normal distribution

```{r}
# Filtering for Euphorbiaceae
euphorbiaceae <- data_trab %>% 
  filter(family == "Euphorbiaceae")
```


```{r}
# QQ-plot for height values
qqnorm(euphorbiaceae$altura)
qqline(euphorbiaceae$altura, col = "red")
```


```{r}
# Normality test
shapiro.test(euphorbiaceae$altura)
```


```{r}
# Estimating parameters
mean_altura <- mean(euphorbiaceae$altura)
sd_altura <- sd(euphorbiaceae$altura)

mean_altura
sd_altura
```


```{r}
# Calculating the empirical probability of height > 20 m among Euphorbiaceae
euphorbiaceae %>% 
  summarise(n_total = n(), n_altura_20 = sum(altura > 20, na.rm = TRUE), prob_empirica = mean(altura > 20, na.rm = TRUE))
```


```{r}
# Calculating the theoretical probability of Euphorbiaceae height > 20 m assuming normality
pnorm(q = 20, mean = mean_altura, sd = sd_altura, lower.tail = FALSE)
```

```{r}
# Calculating the 2.5% and 97.5% quantiles
q_lower <- qnorm(0.025, mean = mean_altura, sd = sd_altura)
q_upper <- qnorm(0.975, mean = mean_altura, sd = sd_altura)

q_lower
q_upper
```


```{r}
# Creating values for the normal curve
normal_curve <- tibble(altura = seq(from = mean_altura - 4 * sd_altura, to = mean_altura + 4 * sd_altura, length.out = 1000), density = dnorm(altura, mean = mean_altura, sd = sd_altura)) %>%
  mutate(tail = case_when(altura <= q_lower ~ "Lower 2.5%", altura >= q_upper ~ "Upper 2.5%", TRUE ~ "Central 95%"))

# Plotting
ggplot(normal_curve, aes(x = altura, y = density)) +
  geom_area(data = filter(normal_curve, tail == "Lower 2.5%"), fill = "red", alpha = 0.8) +
  geom_area(data = filter(normal_curve, tail == "Upper 2.5%"), fill = "red", alpha = 0.8) +
  geom_line(linewidth = 1) +
  geom_vline(xintercept = q_lower, linetype = "dashed") +
  geom_vline(xintercept = q_upper, linetype = "dashed") +
  scale_x_continuous(breaks = c(round(q_lower, 2), round(q_upper, 2))) +
  labs(x = "Height (m)", y = "Density") +
  theme_classic()
```


