---
title: "Inference, hypothesis testing, and sampling"
author: "Jorge Alonso"
date: "2026-05-25"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    code_folding: show
    self_contained: true
---

# 1. Preparing

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 5,
  dpi = 300)
```

## 1.1. Loading packages and data

```{r, message = FALSE}
# Loading packages
library(tidyverse)
library(patchwork)
library(DescTools)
library(infer)
library(car)
```

```{r, message=FALSE, warning=FALSE}
# Cleaning memory
rm(list = ls(all = TRUE))
gc()
```

```{r}
# Database link
processed_url <- "https://raw.githubusercontent.com/jmalonso55/Applied-Statistics-for-a-Tropical-Arboretum/refs/heads/main/data/processed_data.csv"
```


```{r}
# Loading data
data_trab <- read.csv(processed_url, sep = ",")

# Checking
summary(data_trab)
```

## 1.2. Functions

```{r}
# Coefficient of variation function
cv <- function(x) {
  sd(x) / mean(x) * 100
  }
```


```{r}
# Function for descriptive statistics of height
summary_height <- function(data) {
  data %>% summarise(n = n(), mean_height = mean(altura), sd_height = sd(altura), cv_height = cv(altura), median_height = median(altura))
  }
```


```{r}
# Function to compare a sample with the reference population
compare_sample <- function(sample_data, population_data, var) {
  
  pop_values <- population_data %>% pull({{ var }})
  sample_values <- sample_data %>% pull({{ var }})
  
  tibble(statistic = c("mean", "median", "sd", "cv", "q25", "q75", "p90", "p95"),
    population = c(mean(pop_values), median(pop_values), sd(pop_values), cv(pop_values), quantile(pop_values, 0.25), quantile(pop_values, 0.75), quantile(pop_values, 0.90), quantile(pop_values, 0.95)),
    sample = c(mean(sample_values), median(sample_values), sd(sample_values), cv(sample_values), quantile(sample_values, 0.25), quantile(sample_values, 0.75), quantile(sample_values, 0.90), quantile(sample_values, 0.95))) %>%
    mutate(absolute_error = sample - population, relative_error_percent = absolute_error / population * 100)
}
```



# 2. Population and sample

## 2.1. Population parameters: mean, standard deviation, and distribution

```{r}
# Random sample of 50 plants
set.seed(123)
amostra_50 <- data_trab %>%
  slice_sample(n = 50)
```

```{r, fig.width = 8, fig.height = 4}
# Comparing the height distribution in the reference population and in a random sample

# Common x-axis limits
x_lim <- range(data_trab$altura)

# Population histogram
hist_pop <- 
  ggplot(data_trab, aes(x = altura)) +
  geom_histogram(aes(y = after_stat(density)), bins = 10, fill = "#A8D0E6", color = "black") +
  coord_cartesian(xlim = x_lim) +
  labs(x = "Height (m)", y = "Density") +
  scale_y_continuous(expand = c(0, 0)) +
  theme_classic()

# Sample histogram
hist_amo <-  
  ggplot(amostra_50, aes(x = altura)) +
  geom_histogram(aes(y = after_stat(density)), bins = 10, fill = "#A8D0E6", color = "black") +
  coord_cartesian(xlim = x_lim) +
  labs(x = "Height (m)", y = "Density") +
  scale_y_continuous(expand = c(0, 0)) +
  theme_classic()

# Visualizing
hist_pop <- hist_pop + ggtitle("Reference population")
hist_amo <- hist_amo + ggtitle("Random sample, n = 50")

hist_pop + hist_amo
```

```{r}
# Parameter: population mean
# For didactic purposes, the census dataset is treated as the reference population.
m_pop <- mean(data_trab$altura)
paste("Parameter - Population mean =", round(m_pop, 2), "m")


# Estimate: sample mean
# The estimator is the sample mean; the estimate is the value obtained in this sample.
m_amo <- mean(amostra_50$altura)
paste("Estimate - Sample mean =", round(m_amo, 2), "m")


# Sampling error for this specific sample
erro_amostral <- m_amo - m_pop
paste("Sampling error =", round(erro_amostral, 2), "m")
```

```{r}
# Parameter: population standard deviation
sd_pop <- sqrt(sum((data_trab$altura - mean(data_trab$altura))^2) /
    sum(!is.na(data_trab$altura)))
paste("Parameter - Population standard deviation =", round(sd_pop, 2), "m")

# Estimate: sample standard deviation
sd_amo <- sd(amostra_50$altura, na.rm = TRUE)
paste("Estimate - Sample standard deviation =", round(sd_amo, 2), "m")
```

## 2.2. Exploring properties of estimators

```{r}
# Repeated sampling to evaluate the behavior of the sample mean
set.seed(123)

estimativas <- tibble(n_amostral = c(5, 25, 100)) %>% 
  mutate(estimativas = map(n_amostral, ~ replicate(n = 1000, expr = mean(sample(data_trab$altura, size = .x, replace = FALSE))))) %>%
  unnest_longer(estimativas) %>%
  group_by(n_amostral) %>%
  mutate(simulacao = row_number()) %>%
  ungroup()

# Checking the behavior of the estimator
estimativas %>% 
  group_by(n_amostral) %>% 
  reframe(mean_estimate = mean(estimativas), bias = mean(estimativas) - m_pop, sd_estimate = sd(estimativas))
```


```{r}
# Visualizing the sampling distribution of the sample mean
estimativas %>% 
  mutate(n_amostral = factor(n_amostral, levels = c(100, 25, 5), labels = c("n = 100", "n = 25", "n = 5"))) %>%
  ggplot(aes(x = estimativas, color = n_amostral)) +
  geom_density(linewidth = 1) +
  geom_vline(xintercept = m_pop, linetype = "dashed") +
  coord_cartesian(ylim = c(0, 0.6)) +
  labs(x = "Height (m)", y = "Density", color = NULL) +
  theme_classic() +
  theme(legend.position = "bottom",
    legend.background = element_rect(fill = "white", color = NA))
```

As the sample size increases, the sampling distribution of the sample mean becomes more concentrated around the population mean. This illustrates that the sample mean is a consistent estimator of the population mean.

## 2.3. Central Limit Theorem and sampling distributions

```{r}
# Population parameter: proportion of multi-stemmed plants
mul_pop <- mean(data_trab$multicaule)

# Generating repeated random samples
# For a binary variable, the sample mean corresponds to the sample proportion.
set.seed(123)
amostras_mul <- tibble(amostra = 1:1000) %>%
  mutate(dados = map(amostra, ~ data_trab %>% 
                       slice_sample(n = 100, replace = TRUE))) %>%
  unnest(dados)
```


```{r}
# Calculating the sample proportion of multi-stemmed plants in each sample
estimativas_mul <- amostras_mul %>%
  group_by(amostra) %>%
  summarise(n = n(), prop_multicaule = mean(multicaule), sd_multicaule = sd(multicaule), .groups = "drop")

head(estimativas_mul, 10)
```

```{r}
# Summary of the simulated sampling distribution
estimativas_mul %>%
  summarise(mean_of_sample_proportions = mean(prop_multicaule), sd_of_sample_proportions = sd(prop_multicaule), population_proportion = mul_pop)
```

```{r}
# Visualizing the sampling distribution of the sample proportion
ggplot(estimativas_mul, aes(x = prop_multicaule)) +
  geom_histogram(aes(y = after_stat(density)), bins = 24, fill = "#A8D0E6", color = "black") +
  geom_density(linewidth = 1) +
  geom_vline(xintercept = mul_pop, linetype = "dashed", linewidth = 1) +
  labs(x = "Sample proportion of multi-stemmed plants", y = "Density", title = "Sampling distribution of the sample proportion", subtitle = "1,000 samples of 100 plants") +
  theme_classic()
```


Because multicaule is a binary variable, its sample mean represents the sample proportion of multi-stemmed plants. The simulated distribution shows how this proportion varies across repeated samples. According to the Central Limit Theorem, as the sample size increases, the sampling distribution of the sample mean tends to become approximately normal, even when the original variable is binary.


# 3. Confidence intervals

## 3.1. Confidence interval for the mean

```{r}
# Confidence interval for the mean height based on one random sample of 50 plants
ic_50 <- MeanCI(amostra_50$altura, conf.level = 0.95)
ic_50
```

```{r}
# Checking whether the confidence interval contains the population mean
ic_50["lwr.ci"] <= m_pop & ic_50["upr.ci"] >= m_pop
```

## 3.2. Confidence intervals and coverage

```{r}
# Simulating several confidence intervals from repeated samples
set.seed(123)
ic_sim <- tibble(amostra = 1:100) %>%
  mutate(dados = map(amostra, ~ sample(data_trab$altura, size = 50, replace = FALSE)), ic = map(dados, ~ MeanCI(.x, conf.level = 0.95))) %>%
  mutate(media = map_dbl(ic, ~ .x["mean"]), lower = map_dbl(ic, ~ .x["lwr.ci"]), upper = map_dbl(ic, ~ .x["upr.ci"]), contains_mean = lower <= m_pop & upper >= m_pop) %>%
  select(amostra, media, lower, upper, contains_mean)

head(ic_sim)
```


```{r}
# Proportion of confidence intervals that contain the population mean
ic_sim %>%
  summarise(n_intervals = n(), confidence_level = 0.95, coverage = mean(contains_mean))
```


## 3.3. Effect of the confidence level on interval width

```{r}
# Comparing confidence intervals with different confidence levels
# The same random sample is used in all cases.
ic_levels <- tibble(confidence_level = c(0.90, 0.95, 0.99)) %>%
  mutate(ic = map(confidence_level, ~ MeanCI(amostra_50$altura, conf.level = .x)), mean = map_dbl(ic, ~ .x["mean"]), lower = map_dbl(ic, ~ .x["lwr.ci"]), upper = map_dbl(ic, ~ .x["upr.ci"]), width = upper - lower, contains_population_mean = lower <= m_pop & upper >= m_pop) %>%
  select(confidence_level, mean, lower, upper, width, contains_population_mean)

ic_levels
```


```{r}
# Visualizing the effect of the confidence level on interval width
ggplot(ic_levels, aes(x = factor(confidence_level), y = mean)) +
  geom_hline(yintercept = m_pop, linetype = "dashed", linewidth = 1) +
  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.15) +
  geom_point(size = 2) +
  labs(x = "Confidence level", y = "Mean height (m)", title = "Effect of confidence level on interval width", subtitle = "Confidence intervals based on the same random sample of 50 plants") +
  theme_classic()
```


This section illustrates how confidence intervals quantify the uncertainty of sample-based estimates. The census dataset is treated as the reference population, allowing the true population mean to be compared with intervals calculated from random samples. This makes it possible to show that a single confidence interval either contains or does not contain the population mean, whereas the confidence level refers to the long-term proportion of intervals that would contain the parameter under repeated sampling. The section also demonstrates that higher confidence levels produce wider intervals for the same sample because they require a larger margin of error.


# 4. Hypothesis testing

## 4.1. Comparing one group with a reference value from the arboretum

### 4.1.1. Non-parametric test: height in bed 26F

```{r}
# Filtering the 26F bed and the remaining arboretum records
canteiro_26F <- data_trab %>% 
  filter(sec_cant == "26F")

menos_26F <- data_trab %>% 
  filter(sec_cant != "26F")
```


```{r}
# Descriptive statistics for the 26F bed
summary_height(canteiro_26F)
```


```{r}
# Descriptive statistics for the remaining arboretum records
summary_height(menos_26F)
```


```{r}
# Checking normality in the 26F bed
shapiro.test(canteiro_26F$altura)
```

```{r}
# Graphical assessment of normality
qqnorm(canteiro_26F$altura, main = "", xlab = "Theoretical quantiles", ylab = "Sample quantiles")
qqline(canteiro_26F$altura, col = "red", lwd = 2)
```


```{r}
# Testing whether height values in the 26F bed are greater than the median height of the remaining arboretum records
mediana_ref_26F <- median(menos_26F$altura)

wilcox.test(canteiro_26F$altura, mu = mediana_ref_26F, alternative = "greater")
```


### 4.1.2. Parametric test: height in Moraceae


```{r}
# Filtering Moraceae and the remaining arboretum records
moraceae <- data_trab %>% 
  filter(family == "Moraceae")

sem_moraceae <- data_trab %>% 
  filter(family != "Moraceae")
```


```{r}
# Descriptive statistics for Moraceae
summary_height(moraceae)
```


```{r}
# Descriptive statistics for the remaining arboretum records
summary_height(sem_moraceae)
```


```{r}
# Checking normality in Moraceae
shapiro.test(moraceae$altura)
```

```{r}
# Graphical assessment of normality
qqnorm(moraceae$altura, main = "", xlab = "Theoretical quantiles", ylab = "Sample quantiles")
qqline(moraceae$altura, col = "red", lwd = 2)
```


```{r}
# Testing whether the mean height of Moraceae differs from the mean height of the remaining arboretum records
media_ref_moraceae <- mean(sem_moraceae$altura)

t.test(moraceae$altura, mu = media_ref_moraceae, alternative = "two.sided")
```

```{r}
# Aplicando o teste t
t.test(moraceae$altura, mu = 14.22868, alternative = "two.sided")
```

## 4.2. Comparing the mean of two groups

```{r}
# Filtering Euphorbiaceae and Moraceae
data_fam <- data_trab %>% 
  filter(family %in% c("Moraceae", "Euphorbiaceae"))
```


```{r}
# Descriptive statistics by family
data_fam %>% 
  group_by(family) %>% 
  summary_height()
```


```{r}
# Checking taxonomic composition by genus
data_fam %>% 
  count(family, genus, sort = TRUE)
```


```{r}
# Fitting a one-way model for diagnostic purposes
mod_fam <- aov(altura ~ family, data = data_fam)
```


```{r}
# Checking normality of model residuals
shapiro.test(residuals(mod_fam))
```


```{r}
# Graphical assessment of residual normality
qqnorm(residuals(mod_fam), main = "", xlab = "Theoretical quantiles", ylab = "Sample quantiles")
qqline(residuals(mod_fam), col = "red", lwd = 2)
```


```{r}
# Checking homogeneity of variances
var.test(altura ~ family, data = data_fam)
```


```{r}
# Comparing mean height between families
t.test(altura ~ family, data = data_fam, alternative = "two.sided", var.equal = TRUE)
```


```{r}
# The same two-group comparison can also be expressed as a one-way ANOVA
summary(mod_fam)
```


This section compares mean height between two botanical families. The example illustrates the use of a two-sample Student’s t-test and the evaluation of model assumptions based on residual normality and variance homogeneity. Because the homogeneity of variances assumption was considered acceptable, the t-test was performed assuming equal variances. A one-way ANOVA with two groups is also shown to illustrate that, in this case, ANOVA and the two-sample t-test address the same general hypothesis of equal group means.


## 4.3. Comparing the proportion of multi-stemmed individuals between two families

```{r}
# Creating a table with the variables of interest
prop_family <- data_fam %>%
  group_by(family) %>%
  summarise(n_total = n(), n_multistemmed = sum(multicaule == 1), prop_multistemmed = n_multistemmed / n_total, .groups = "drop") %>%
  arrange(desc(n_total))

prop_family
```

```{r}
# Comparing the proportion of multi-stemmed individuals between families
prop.test(x = prop_family$n_multistemmed, n = prop_family$n_total, alternative = "two.sided")
```


This section compares the proportion of multi-stemmed individuals between two botanical families. Because the response variable is binary, the analysis is based on counts of individuals with and without the attribute in each group. The proportion test evaluates whether the observed difference between family-level proportions is greater than expected under the null hypothesis of equal proportions.


## 4.4. Bootstrap confidence interval and permutation test


This section uses simulation-based inference to compare mean height between multi-stemmed and single stem individuals. First, bootstrap resampling is used to estimate the uncertainty associated with the observed difference in means and to construct a confidence interval. Then, a permutation test is used to evaluate the null hypothesis that height and multistem status are independent.


```{r}
# Observed difference in mean height
# The order c("No", "Yes") calculates mean height of non-bifurcated
# individuals minus mean height of bifurcated individuals.
d_hat <- data_trab %>% 
  specify(altura ~ multicaule_l) %>% 
  calculate(stat = "diff in means", order = c("No", "Yes"))

d_hat
```

```{r}
# Bootstrap distribution of the difference in means
set.seed(333)
boot_bif <- data_trab %>% 
  specify(altura ~ multicaule_l) %>% 
  generate(reps = 10000, type = "bootstrap") %>% 
  calculate(stat = "diff in means", order = c("No", "Yes"))

# Bootstrap confidence interval
ci_boot_bif <- boot_bif %>% 
  get_confidence_interval(level = 0.95, type = "percentile")

ci_boot_bif
```

```{r}
# Visualizing the bootstrap confidence interval
boot_bif %>%
  visualise() +
  shade_confidence_interval(ci_boot_bif)
```


```{r, warning = F}
# Permutation test for the null hypothesis of independence between height and bifurcation status
set.seed(333)
perm_bif <- data_trab %>% 
  specify(altura ~ multicaule_l) %>% 
  hypothesize(null = "independence") %>%
  generate(reps = 10000, type = "permute") %>%
  calculate(stat = "diff in means", order = c("No", "Yes"))

# Permutation p-value
p_perm_bif <- perm_bif %>%
  get_p_value(obs_stat = d_hat, direction = "two-sided")


# p-value
p_val <- p_perm_bif$p_value

if (p_val == 0) {
  paste("p <", 1 / 10000)
} else {
  paste("p =", round(p_val, 4))
}
```


```{r}
# Visualizing the permutation test
perm_bif %>%
  visualise() +
  shade_p_value(obs_stat = d_hat, direction = "two-sided")
```


The bootstrap distribution represents the uncertainty around the observed difference in mean height between the two groups. If the confidence interval does not include zero, this suggests a difference in mean height between multi-stemmed and single stem individuals. The permutation test evaluates the same question under the null hypothesis of independence between height and multistem status. Small p-values indicate that the observed difference would be unlikely if multistem status and height were unrelated.


# 5. Sampling

```{r}
# General settings
z <- 1.96
relative_error <- 0.05
N <- nrow(data_trab)
```


## 5.1. Sample size based on relative error

```{r}
# Sample size based on relative error
n_dap <- (z^2 * (sd(data_trab$dap) / mean(data_trab$dap))^2) / relative_error^2
n_alt <- (z^2 * (sd(data_trab$altura) / mean(data_trab$altura))^2) / relative_error^2

n_dap
n_alt
```


```{r}
# Finite population correction
n_dap_f <- n_dap / (1 + ((n_dap - 1) / N))
n_alt_f <- n_alt / (1 + ((n_alt - 1) / N))

n_dap_f
n_alt_f
```


## 5.2. Simple random sampling

```{r}
# Simple random sampling
set.seed(333)
amo_aleatoria <- data_trab %>% 
  slice_sample(n = round(n_dap, 0), replace = FALSE)

nrow(amo_aleatoria)
```

```{r}
# Comparing sample estimates with population parameters
compare_sample(amo_aleatoria, data_trab, dap)
```


## 5.3. Systematic sampling

```{r}
# Systematic sampling
set.seed(333)

k <- N / round(n_dap, 0)

inicio <- runif(1, min = 1, max = k)

indices_sistematicos <- floor(inicio + (0:(round(n_dap, 0) - 1)) * k)

indices_sistematicos <- indices_sistematicos[indices_sistematicos <= N]

amo_sistematica <- data_trab[indices_sistematicos, ]

nrow(amo_sistematica)
```

```{r}
# Comparing sample estimates with population parameters
compare_sample(amo_sistematica, data_trab, dap)
```


## 5.4. Stratified sampling

```{r}
# Proportional stratified sampling by bed
set.seed(123)

alocacao <- data_trab %>%
  filter(sec_cant != "") %>%
  count(sec_cant, name = "N_h") %>%
  mutate(prop = N_h / sum(N_h), n_h_raw = prop * round(n_dap, 0), n_h = floor(n_h_raw), remainder = n_h_raw - n_h)

# Adjusting allocation to obtain exactly the target sample size
faltam <- round(n_dap, 0) - sum(alocacao$n_h)

alocacao <- alocacao %>%
  arrange(desc(remainder)) %>%
  mutate(n_h = n_h + if_else(row_number() <= faltam, 1L, 0L)) %>%
  select(sec_cant, N_h, n_h)

# Drawing the proportional stratified sample
amo_estratificada <- data_trab %>%
  filter(sec_cant != "") %>%
  inner_join(alocacao, by = "sec_cant") %>%
  group_by(sec_cant) %>%
  group_modify(~ slice_sample(.x, n = unique(.x$n_h), replace = FALSE)) %>%
  ungroup() %>%
  select(-N_h, -n_h)

nrow(amo_estratificada)
```

```{r}
# Comparing sample estimates with population parameters
compare_sample(amo_estratificada, data_trab, dap)
```


## 5.5. Cluster sampling

### 5.5.1. One-stage cluster sampling

```{r}
# One-stage cluster sampling
# Entire beds are selected until the accumulated number of individuals reaches the target sample size.
set.seed(123)

conglomerados <- data_trab %>%
  filter(sec_cant != "") %>%
  count(sec_cant, name = "N_cant")

conglomerados_sorteados <- conglomerados %>%
  slice_sample(prop = 1) %>%
  mutate(N_acum = cumsum(N_cant)) %>%
  filter(N_acum <= round(n_dap, 0) | lag(N_acum, default = 0) < round(n_dap, 0))

amo_conglomerados <- data_trab %>%
  filter(sec_cant %in% conglomerados_sorteados$sec_cant)

nrow(amo_conglomerados)
```

```{r}
# Comparing sample estimates with population parameters
compare_sample(amo_conglomerados, data_trab, dap)
```


### 5.5.2. Two-stage cluster sampling

```{r}
# Two-stage cluster sampling
# First, beds are selected. Then, individuals are sampled within the selected beds.
set.seed(123)

n_canteiros <- 50 

canteiros_sorteados <- data_trab %>%
  filter(sec_cant != "") %>%
  distinct(sec_cant) %>%
  slice_sample(n = n_canteiros)

base_canteiros <- data_trab %>%
  filter(sec_cant %in% canteiros_sorteados$sec_cant)

amo_conglomerados_2est <- base_canteiros %>%
  slice_sample(n = round(n_dap, 0), replace = FALSE)

nrow(amo_conglomerados_2est)
```

```{r}
# Comparing sample estimates with population parameters
compare_sample(amo_conglomerados_2est, data_trab, dap)
```


## 5.6. Comparing sampling strategies


```{r}
# Organizing sampling outputs in a named list
samples_list <- list(
  "Simple random sampling" = amo_aleatoria,
  "Systematic sampling" = amo_sistematica,
  "Stratified sampling" = amo_estratificada,
  "One-stage cluster sampling" = amo_conglomerados,
  "Two-stage cluster sampling" = amo_conglomerados_2est
)
```


```{r}
# Comparing each sample with the reference population
sampling_comparison <- imap_dfr(
  samples_list,
  ~ compare_sample(.x, data_trab, dap) %>%
    mutate(method = .y)
) %>%
  relocate(method)

sampling_comparison
```


```{r}
# Plotting relative error of mean DBH
sampling_comparison %>%
  filter(statistic == "mean") %>%
  mutate(method = reorder(method, abs(relative_error_percent))) %>%
  ggplot(aes(x = relative_error_percent, y = method)) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  geom_vline(xintercept = c(-5, 5), linetype = "dotted") +
  geom_segment(aes(x = 0, xend = relative_error_percent, yend = method), linewidth = 0.8) +
  geom_point(size = 3) +
  geom_text(aes(label = paste0(round(relative_error_percent, 2), "%"), hjust = if_else(relative_error_percent >= 0, -0.2, 1.2)), size = 3.5) +
  scale_x_continuous(limits = c(-6, 7), breaks = c(-5, -2.5, 0, 2.5, 5)) +
  labs(x = "Relative error of mean DBH (%)", y = "Sampling method", title = "Relative error of mean DBH by sampling method", subtitle = "Dotted lines indicate the ±5% relative error threshold") +
  theme_classic()
```


This section introduces different sampling strategies using the arboretum census as the reference population. First, sample size is estimated based on the desired relative error for DBH and height. Then, different sampling designs are applied, including simple random sampling, systematic sampling, stratified sampling, and cluster sampling. For each design, the sample mean and standard deviation of DBH are compared with the corresponding values calculated from the reference population. The objective is not to define a definitive sampling protocol, but to illustrate how different sampling designs can produce different estimates from the same population.

