---
title: "Trees regression"
author: "Jorge Alonso"
date: "2026-05-30"
output: 
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    code_folding: show
    self_contained: true
---

# 1. Preparing for analysis

## 1.1. Loading data and packages

```{r}
# Loading packages
library(tidyverse)
library(DHARMa)
library(car)
library(lmtest)
library(MASS)
library(lme4)
library(lmerTest)
library(performance)
library(emmeans)
```

```{r}
# Limpeza do histórico e da memória
rm(list=ls(all=TRUE)) 
gc() 
```

```{r}
# Database link
processed_url <- "https://raw.githubusercontent.com/jmalonso55/Applied-Statistics-for-a-Tropical-Arboretum/refs/heads/main/data/processed_data.csv"
```


```{r}
# Loading data
data_trab <- read.csv(processed_url, sep = ",")

# Checking
summary(data_trab)
```

## 1.2. Filtering database

```{r}
# Filtering for trees
arvore <- data_trab %>% 
  filter(habito_pad == "Tree")
```


```{r}
# Checking family
sort(unique(arvore$family))
```


```{r}
# Removing NA and tranforming variables
arvore_mod <- arvore %>%
  filter(!is.na(altura),
    !is.na(dap),
    !is.na(n_caule),
    !is.na(multicaule),
    !is.na(sec_cant),
    !is.na(family)) %>%
  mutate(sec_cant = as.factor(sec_cant),
    family = as.factor(family),
    multicaule = as.factor(multicaule))
```

```{r}
# Filtering the database for family represented by 10 or more plants
arvore_mod2 <- arvore_mod %>%
  filter(!is.na(family)) %>%
  add_count(family, name = "n_family") %>%
  filter(n_family >= 10) %>%
  dplyr::select(-n_family)
```


## 1.3. Height x DHB visualization

```{r}
# Counting 16 top families
contagem <- arvore_mod2 %>% 
  filter(!is.na(family)) %>% 
  count(family) %>% 
  arrange(desc(n)) %>% 
  slice_head(n = 16) %>% 
  pull(family)

# Filtering
arv_16 <- arvore_mod2 %>% 
  filter(family %in% contagem) 
```


```{r}
# Visualizing
ggplot(arv_16, aes(x = dap, y = altura, color = multicaule_l)) +
  geom_point(alpha = 0.4) +
  facet_wrap(~family) +
  theme_classic()
```

# 2. Modeling for specific genus/families

## 2.1. Meliaceae

```{r}
# Filtering for genus and beds with more than 3 plants
meliaceae <- arvore %>%
  filter(family == "Meliaceae") %>% 
  filter(!is.na(genus)) %>%
  add_count(genus, name = "n_genus") %>%
  filter(n_genus >= 3) %>%
  add_count(sec_cant, name = "n_sec_cant") %>%
  filter(n_sec_cant >= 3) %>%
  dplyr::select(-n_genus, -n_sec_cant)
```


### 2.1.1. Linear model

```{r}
# Linear model
mod1 <- lm(altura ~ dap + genus, data = meliaceae)
```

```{r}
# Analysis of residuals
res_mod1 <- simulateResiduals(mod1, n = 1000)
plot(res_mod1)
```

```{r}
# Linear model ANOVA
anova(mod1)
```

```{r}
# Linear model summary
summary(mod1)
```

### 2.1.2. Log model

```{r}
# Log model
mod_log <- lm(altura ~ log(dap) + genus, data = meliaceae)
```

```{r}
# Analysis of residuals
res_mod_log <- simulateResiduals(mod_log, n = 1000)
plot(res_mod_log)
```

### 2.1.3. Quadratic model

```{r}
# Quadratic model
mod2 <- lm(altura ~ dap + I(dap^2) + genus, data = meliaceae)
```


```{r}
# Analysis of residuals
res_mod2 <- simulateResiduals(mod2, n = 1000)
plot(res_mod2)
```

```{r}
# Quadratic model ANOVA
anova(mod2)
```

### 2.1.4. Comparing models

```{r}
# ANOVA for comparing linear, quadratic and log regression models
anova(mod1, mod2, mod_log)
```


```{r}
# AIC and BIC for model comparison
AIC(mod1, mod2, mod_log)
BIC(mod1, mod2, mod_log)
```

```{r}
# Linear model R²
r2(mod1)
```

```{r}
# log model R²
r2(mod_log)
```

### 2.1.5. Visualizing results

```{r}
# Means comparison between species
emm_sp <- emmeans(mod_log, ~ genus)

multcomp::cld(emm_sp, adjust = "sidak", Letters = letters, decreasing = T)
```

```{r}
ggplot(data = meliaceae, aes(x = dap, y = altura)) +
  geom_smooth(method = "lm", formula = y ~ log(x)) +
  geom_point(aes(color = genus)) +
  theme_classic()
```

## 2.2. Carapa - Meliaceae genus

```{r}
carapa <- arvore %>%
  filter(genus == "Carapa")
```

### 2.2.1. Linear model

```{r}
# Linear model
mod1c <- lm(altura ~ dap, data = carapa)
```

```{r}
# Analysis of residuals
res_mod1c <- simulateResiduals(mod1c, n = 1000)
plot(res_mod1c)
```

### 2.2.2. Log model

```{r}
# Log model
mod_logc <- lm(altura ~ log(dap), data = carapa)
```

```{r}
# Analysis of residuals
res_mod_logc <- simulateResiduals(mod_logc, n = 1000)
plot(res_mod_logc)
```

```{r}
# log model ANOVA
anova(mod_logc)
```

```{r}
# log model summary
summary(mod_logc)
```

### 2.2.3. Quadratic model

```{r}
# Quadratic model
mod2c <- lm(altura ~ dap + I(dap^2), data = carapa)
```


```{r}
# Analysis of residuals
res_mod2c <- simulateResiduals(mod2c, n = 1000)
plot(res_mod2c)
```

### 2.2.4. Mixed effects log model

```{r}
# log model with beds random effects
mod1m <- lmer(altura ~ log(dap) + (1 | sec_cant), data = carapa, REML = T)
```


```{r}
# Analysis of residuals
res_mod1m <- simulateResiduals(mod1m, n = 1000)
plot(res_mod1m)
```


```{r}
# Comparing models without and with random effects
AIC(mod_logc, mod1m)
BIC(mod_logc, mod1m)
```

```{r}
# log model R²
r2(mod_logc)
```

```{r}
# mixed model R²
r2(mod1m)
```


```{r}
# Mixed model summary
summary(mod1m)
```


```{r}
# Mixed model summary
anova(mod1m)
```

```{r}
# Random effects significance
ranova(mod1m)
```

### 2.2.5. Results visualization

```{r}
# Predicting for plotting
pred_mod1m <- ggpredict(mod1m, terms = "dap [all]", type = "fixed")

# Visualizing
ggplot(pred_mod1m, aes(x = x, y = predicted)) +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.2) +
  geom_line(linewidth = 1.2) +
  geom_point(data = carapa, aes(x = dap, y = altura, color = sec_cant), inherit.aes = FALSE, alpha = 0.7) +
  labs(x = "DAP", y = "Altura predita", title = "") +
  theme_classic()
```


```{r}
# Extraindo efeitos aleatórios dos canteiros
rank_canteiros <- ranef(mod1m)$sec_cant %>%
  as.data.frame() %>%
  rownames_to_column("sec_cant") %>%
  rename(efeito_canteiro = `(Intercept)`) %>%
  arrange(desc(efeito_canteiro)) %>%
  mutate(rank = row_number(), interpretacao = case_when(
      efeito_canteiro > 0 ~ "Acima da média",
      efeito_canteiro < 0 ~ "Abaixo da média",
      TRUE ~ "Na média"))
```


```{r}
# Visualizando
ggplot(rank_canteiros, aes(x = reorder(sec_cant, efeito_canteiro), y = efeito_canteiro)) +
  geom_col() +
  geom_hline(yintercept = 0, linetype = "dashed") +
  coord_flip() +
  labs(x = "Canteiro", y = "Efeito aleatório estimado", title = "Ranking dos canteiros pelo efeito aleatório") +
  theme_classic()
```


## 2.3. Bignoniaceae

```{r}
# Filtering database
bignoniaceae <- arvore %>%
  filter(family == "Bignoniaceae", !is.na(genus), genus != "", !is.na(sec_cant), sec_cant != "") %>%
  add_count(genus, name = "n_genus") %>%
  filter(n_genus >= 3) %>%
  dplyr::select(-n_genus) %>%
  add_count(sec_cant, name = "n_sec_cant") %>%
  filter(n_sec_cant >= 4) %>%
  dplyr::select(-n_sec_cant) %>%
  add_count(genus, name = "n_genus_final") %>%
  filter(n_genus_final >= 3) %>%
  dplyr::select(-n_genus_final)
```


### 2.3.1. Linear model

```{r}
# Linear model
mod1b <- lm(altura ~ dap + genus, data = bignoniaceae)
```

```{r}
# Analysis of residuals
res_mod1b <- simulateResiduals(mod1b, n = 1000)
plot(res_mod1b)
```

```{r}
# Model ANOVA
anova(mod1b)
```

```{r}
# Linear model without genus
mod1b_alt <- lm(altura ~ dap, data = bignoniaceae)
```


```{r}
# Analysis of residuals
res_mod1b_alt <- simulateResiduals(mod1b_alt, n = 1000)
plot(res_mod1b_alt)
```


```{r}
# Simple linear model summary
summary(mod1b_alt)
```

### 2.3.2. Log model

```{r}
# Log model
mod_logb <- lm(altura ~ log(dap) + genus, data = bignoniaceae)
```


```{r}
# Analysis of residuals
res_mod_logb <- simulateResiduals(mod_logb, n = 1000)
plot(res_mod_logb)
```

```{r}
# log model ANOVA
anova(mod_logb)
```

```{r}
# Simple log model
mod_logb_alt <- lm(altura ~ log(dap), data = bignoniaceae)
```

```{r}
# Analysis of residuals
res_mod_logb_alt <- simulateResiduals(mod_logb_alt, n = 1000)
plot(res_mod_logb_alt)
```


```{r}
# Simple log summary
summary(mod_logb_alt)
```


### 2.3.3. Quadratic model

```{r}
# Quadratic model
mod2b <- lm(altura ~ dap + I(dap^2) + genus, data = bignoniaceae)
```


```{r}
# Analysis of residuals
res_mod2b <- simulateResiduals(mod2b, n = 1000)
plot(res_mod2b)
```

```{r}
# Quadratic model ANOVA
anova(mod2b)
```



```{r}
# Simple quadratic model
mod2b_alt <- lm(altura ~ dap + I(dap^2), data = bignoniaceae)
```

```{r}
# Analysis of residuals
res_mod2b_alt <- simulateResiduals(mod2b_alt, n = 1000)
plot(res_mod2b_alt)
```

### 2.3.4. Model comparison

```{r}
# ANOVA for comparing log and quadratic regression models
anova(mod_logb_alt, mod2b_alt)
```

```{r}
# Model comparison
AIC(mod_logb_alt, mod2b_alt)
BIC(mod_logb_alt, mod2b_alt)
```

```{r}
# log model R²
r2(mod_logb_alt)
```

```{r}
# Quadratic model R²
r2(mod2b_alt)
```

### 2.3.5. Mixed effects log model

```{r}
# Model with beds as random effects
mod1mb <- lmer(altura ~ log(dap) + (1 | sec_cant), data = bignoniaceae)
```

```{r}
# Analysis of residuals
res_mod1mb <- simulateResiduals(mod1mb, n = 1000)
plot(res_mod1mb)
```

```{r}
# Model ANOVA
anova(mod1mb)
```

```{r}
# Model R²
r2(mod1mb)
```

```{r}
# Random effects significance
ranova(mod1mb)
```

```{r}
# Model comparison
AIC(mod_logb_alt, mod1mb)
BIC(mod_logb_alt, mod1mb)
```

### 2.3.6. Results visualization

```{r}
# Predicting for plotting
pred_mod1mb <- ggpredict(mod1mb, terms = "dap [all]", type = "fixed")

# Visualizing
ggplot(pred_mod1mb, aes(x = x, y = predicted)) +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.2) +
  geom_line(linewidth = 1.2) +
  geom_point(data = bignoniaceae, aes(x = dap, y = altura, color = sec_cant), inherit.aes = FALSE, alpha = 0.7) +
  labs(x = "DAP", y = "Altura predita", title = "") +
  theme_classic()
```


```{r}
# Extraindo efeitos aleatórios dos canteiros
rank_canteiros_b <- ranef(mod1mb)$sec_cant %>%
  as.data.frame() %>%
  rownames_to_column("sec_cant") %>%
  rename(efeito_canteiro = `(Intercept)`) %>%
  arrange(desc(efeito_canteiro)) %>%
  mutate(rank = row_number(), interpretacao = case_when(efeito_canteiro > 0 ~ "Acima da média",
      efeito_canteiro < 0 ~ "Abaixo da média",
      TRUE ~ "Na média"))
```


```{r}
# Visualizando
ggplot(rank_canteiros_b, aes(x = reorder(sec_cant, efeito_canteiro), y = efeito_canteiro)) +
  geom_col() +
  geom_hline(yintercept = 0, linetype = "dashed") +
  coord_flip() +
  labs(x = "Canteiro", y = "Efeito aleatório estimado", title = "Ranking dos canteiros pelo efeito aleatório") +
  theme_classic()
```

## 2.4. Binomial model for multicaule in Fabaceae

```{r}
# Filtering the database
fabaceae  <- arvore %>%
  filter(family == "Fabaceae", !is.na(genus), genus != "", !is.na(sec_cant), sec_cant != "") %>%
  add_count(genus, name = "n_genus") %>%
  filter(n_genus >= 3) %>%
  dplyr::select(-n_genus) %>%
  add_count(sec_cant, name = "n_sec_cant") %>%
  filter(n_sec_cant >= 3) %>%
  dplyr::select(-n_sec_cant) %>%
  add_count(genus, name = "n_genus_final") %>%
  filter(n_genus_final >= 5) %>%
  dplyr::select(-n_genus_final)
```



```{r}
# Binomial model
mod_binom <- glm(multicaule ~ altura + genus, data = fabaceae, family = "binomial")
```

```{r}
# Analysis of residuals
res_mod_binom <- simulateResiduals(mod_binom, n = 1000)
plot(res_mod_binom)
```


```{r}
# Model ANOVA
anova(mod_binom)
```

```{r}
# Model summary
summary(mod_binom)
```

```{r}
# Predição média da probabilidade de multicaule em função da altura
pred_altura <- ggpredict(mod_binom, terms = "altura [all]")

# Visualizing
ggplot() +
  geom_jitter(data = fabaceae, aes(x = altura, y = multicaule), height = 0.05, width = 0.05, alpha = 0.35) +
  geom_ribbon(data = pred_altura, aes(x = x, ymin = conf.low, ymax = conf.high), alpha = 0.2) +
  geom_line(data = pred_altura, aes(x = x, y = predicted), linewidth = 1.2) +
  scale_y_continuous(limits = c(-0.1, 1.1), breaks = seq(0, 1, 0.2)) +
  labs(x = "Altura", y = "Probabilidade predita de multicaule") +
  theme_classic()
```


```{r}
# Predicting considering genus
pred_genus <- ggpredict(mod_binom, terms = "genus")

# Visualizing
ggplot(pred_genus, aes(x = reorder(x, predicted), y = predicted)) +
  geom_pointrange(aes(ymin = conf.low, ymax = conf.high)) +
  coord_flip() +
  scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.2)) +
  labs(x = "Gênero", y = "Probabilidade predita de multicaule") +
  theme_classic()
```


